---
title: Performance of individual audit filters in predicting opportunities for improvement
  in adult trauma patients
author: "Josef Al-Khalili"
subtitle: A single-center registry-based retrospective cohort
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: 
    reference_docx: JosefAlKhaliliExJobbSeminarieVersion.docx
csl: bmcemerg.csl
editor_options:
  markdown:
    wrap: 72
bibliography: bibliography.bib
always_allow_html: yes
---

```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
source("main.R")
library(rofi)
library(dplyr)
library(ggplot2)
library(pROC)
library(gtsummary)
library(gt)
library(boot)
library(webshot2)
library(knitr)
```

# Abbreviations

OFI -- Opportunities for Improvement

CI -- Confidence Interval

GCS -- Glasgow Coma Scale

ISS -- Injury Severity Score

ED -- Emergency Department

AUC -- Area Under Curve

ROC -- Receiver Operating Characteristics

CPR -- Cardiopulmonary Resuscitation

SBP -- Systolic Blood Pressure

WHO -- World Health Organisation

YLL -- Years Life Lost

DALY -- Disability Adjusted Life Years

PTSD -- Post Traumatic Stress Disorder

# Sammanfattning

*Bakgrund:* Trauma står för 9% av globala dödsfall årligen och har en
betydande påverkan på individer och samhället.
Kvalitetsförbättringsprogram (QI) syftar till att minimera dessa
ogynnsamma utfall genom användning av audit filter som indikatorer för
standardvård och genom att utvärdera patientfall för potentiella
förbättringsmöjligheter (OFI). *Syfte:* Denna studie syftade till att
utvärdera prestandan hos audit filter för att förutsäga OFI och att
bedöma deras relevans som screeningverktyg för Morbiditets- och
Mortalitetskonferenser (MMC). *Material och Metoder:* Sensitivitet,
specificitet och Area Under the Receiver Operating Characteristics Curve
(AUC) beräknades för att utvärdera prestanda. En t-test för AUC
genomfördes för att jämföra prestanda med slumpmässighet. *Resultat:* Av
de 11 audit filter som utvärderades hade 3 en AUC på 0,5 (p \> 0,05). De
återstående audit filterna visade en korrelation med OFI, med AUC-värden
som varierade från 0,51 till 0,63. Audit filtret "\> 60 min till första
ingrepp" uppvisade den högsta AUC på 0,63. *Slutsats:* Ingen enskild
revisionsfilter tycktes prestera väl i att förutsäga OFI. Detta antyder
behovet av att ompröva den nuvarande screeningmetoden för MMC.

# Abstract

*Introduction:* Trauma is responsible for 9% of global deaths anually
and has a significant impact on individuals and society. Quality
Improvement (QI) programs seek to minimize these adverse outcomes
through the use of audit filters as indicators of standard care and
through assessing patient cases for potential Opportunities for
Improvement (OFI). *Aims:* This study aimed to evaluate the performance
of audit filters in predicting OFI and to assess their relevance as a
screening tool for Morbidity and Mortality Conferences (MMC). *Material
and Methods:* The sensitivity, specificity and Area Under the Reciever
Operating Characteristics Curve (AUC) was calculate din order to assess
performance. A t-test for AUC was conducted to compare the performance
to coincidence. *Results:* Of the 11 audit filters assessed, 3 had an
AUC of 0.5 (p \> 0.05). The remaining audit filters demonstrated a
correlation with OFI, with AUC ranging from 0.51 to 0.63. The audit
filter "\> 60 min until first intervention" exhibited the highest AUC of
0.63. *Conclusion:* No individual audit filter seemed to perform well in
predicting OFI, which suggest rethinking the current screening method
for MMC.

*Keywords:* Trauma, Opportunities for Improvement, audit filter,
performance

# Introduction

Trauma is defined as a physical injury of external origin together with
the associated reaction of the body. This phenomenon accounts for 8-9% of
global deaths annually [@world2014injuries; @roth2018global]. Trauma affects low- and
middle-income countries at a higher rate compared to high-income
countries. In addition, young people are at higher risk of trauma
compared to the general population. It is the leading cause of death in
people between the ages of 10 and 49 [@vos2020global; @roth2018global].
The most common type of trauma for people of all ages is road traffic
injuries [@vos2020global; @roth2018global]. In Sweden, road traffic
injuries account for approximately 40% of injuries and falls for 40%
[@SweTrau2022Arsrapport].

Trauma is broadly divided into two groups according to the underlying
causal mechanism: penetrating and blunt trauma. Commonly known examples
of penetrating trauma are gunshot wounds and knife injuries. Penetrating trauma is often related to interpersonal violence or attempted suicide [@stormann2016epidemiology]. Blunt
trauma, on the other hand, is caused by injuries such as road traffic
accidents and falls [@stormann2016epidemiology]. In general, both types of trauma can also be
characterised by the magnitude of the force causing the injury and the
area of impact. Blunt trauma is caused by a high force impact over a
larger area. In contrast, penetrating trauma is caused by a force
applied to a smaller area, causing greater pressure so that it is
penetrating and often results in deeper lacerations to tissues and
organs. However, blunt trauma with sufficient force can be penetrating
in nature [@hexdall2011Mechanism]. Trauma varies widely and has both
long and short term consequences on an individual, community, societal and economic
level [@newnam2014impacts].

Trauma is responsible for 11.9% of years of life lost (YLL) worldwide
[@roth2018global] and the top cause for disability-adjusted life-years (DALY) [@vos2020global]. In addition, trauma is associated with a higher
propensity towards depression, post-traumatic stress disorder (PTSD) and
negative effects on cognitive function [@kessler1995posttraumatic; @north1999psychological;
@muscatelli2017prevalence; @newnam2014impacts]. It is estimated that up to 11% of trauma patients have PTSD three years
after a traumatic event [@visser2017course]. In addition, people who
witness trauma are also at a higher risk of developing PTSD. Trauma is
associated with 36% not returning to work within one year
[@david2022measuring].

However, trauma is not only a burden on the person affected, but also on
communities and society [@newnam2014impacts]. One aspect of such societal
impact is the supposed high economic cost that comes with trauma [@wesson2014cost]. Attempts have been made to estimate the
global cost of trauma, which is a complicated task due to its
variability. However, it is estimated that the cost of road traffic
injuries alone accounts for 2% of gross domestic product (GDP) in
high-income countries [@world2014injuries].

## Trauma Care Systems

Trauma care is a multidisciplinary, complex and time-critical health
care service provided by hospitals. A variety of different specialities
are involved in trauma care, including surgeons, orthopaedists,
anaesthesiologists, emergency physicians and nurses [@American2022resources] Trauma care is
divided into three components: pre-hospital, in-hospital and
post-hospital. The pre-hospital component constitutes first responders,
basic life support and advanced life support [@world2005prehospital].
The in-hospital component is defined as the moment a trauma patient is
admitted to a health care facility, often a trauma centre. The
post-hospital component involves the rehabilitation and recovery of
treated trauma patients.

Trauma centres have a significantly lower mortality rate compared to
non-trauma centres, and the maturity of a trauma centre correlates with
lower mortality, demonstrating the importance of high-quality,
specialised trauma care [@alharbi2021effectiveness;
@mann1999systematic; @candefjord2020mortality]. Despite this, even
established trauma care systems are prone to errors due to the complex nature of trauma and its susceptibility to delays [@nikouline2021errors;
@vioque2014classifying; @o2013opportunities], which in turn can lead to severe or even life-threatening complications. Studies between 1990 and 2014 have shown a pooled preventable
death rate of 20%, with more recent studies showing a lower preventable death rate
[@kwon2014pooled]. For example, a recent study in a Swedish Level I
trauma centre found a preventable death rate of 4%
[@ghorbani2018analysis]. 

## Quality Improvement Programmes

To minimise errors in trauma care, quality improvement (QI) programs
have been established by the World Health Organization (WHO) and the
International Association for Trauma Surgery and Intensive Care (IATSIC)
[@world2009guidelines]. These programs form the core of current trauma
care systems and their development around the world [@American2022resources].

QI programs consist of a series of recommended techniques aiming to
facilitate discussion and analysis of the trauma process in order to
take corrective action for future success. The following techniques are commonly
used: a multidisciplinary mortality and morbidity conference (MMC), a
preventable death review panel, and the use of audit
filters also known as quality indicators [@world2009guidelines].

### Morbidity and Mortality Conferences and Preventable Death Review Panels

The MMC is a central QI technique. It is a proven framework with the
critical components of anonymity, focusing on specific adverse events,
recognising flawed approaches, critical analysis, improvement proposals
and implementing such [@gregor2016morbidity].

QI-centred MMCs are regular meetings broadly defined by five themes: (1) a
clear definition of the role of the MMC, (2) involvement of
stakeholders, (3) recognising and selecting appropriate patient cases
for presentation, (4) structured and focused discussions, and (5)
formulating recommendations and assigning follow-up actions
[@giesbrecht2016morbidity; @churchill2020quality].

The conference participants are healthcare professionals
involved in trauma care and its processes, but not necessarily the
specific patient case. The selection of patient cases varies, but common
themes for discussion are unexpected mortality, unexpected morbidity and
errors [@orlander2003morbidity]. It is recommended that they last about
40 minutes, but this varies between hospitals
[@world2009guidelines].

Implemented effectively, MMCs serve as an important tool for error
reduction without being perceived as negative feedback for individual
care providers, but rather as an incentive for reporting. These
conferences promote a more comprehensive, systems-based approach that
emphasises task analysis, teamwork and quality improvement
[@lazzara2022morbidity]. The introduction of MMC has been shown to
increase morbidity and mortality reporting [@kong2016analysis].

### Audit Filters and Opportunities for Improvement

Audit filters, also known as quality indicators are established
criteria based on standardised care. Any deviation from the audit filters should raise the attention of health care professionals with the purpose of improving the
quality of care [@American2022resources]. An audit filter can involve
outcomes such as "death" or criteria related to guidelines and gold
standards such as "the placement of two large bore intravenous lines
within 15 minutes from arrival to a health care facility"
[@world2009guidelines].   

Previous studies have shown a reduction in trauma-related mortality
after the implementation of audit filters
[@chadbunchachai2001comparative; @chadbunchachai2003study]. Per contra, newer studies seem to
differ. In 2009, an extensive review by Evans et al. [@evans2009audit] sought to determine if audit
filters could be used in improving processes in trauma care. They found no studies successfully providing evidence for the use of audit filters. More than 1,500 audit filters have been identified in a review by Stelfox et al. [@stelfox2010quality]. Notably, they did not find a common set of clearly defined, evidence-based, broadly accepted audit filters for evaluating the quality of trauma care. The perceived usefulness of audit filters still remains high [@berg2022perceived]   

Opportunities for improvement (OFIs) are the endpoint of patient care.
Following the review of individual patient cases there is a
decision made regarding the existence of an OFI [@American2022resources]. It has been shown that
such review processes are associated with high-quality trauma care
[@santana2014development]. OFIs are typically associated with failures
in initial care [@vioque2014classifying], specifically in airway
management, fluid resuscitation, haemorrhage control and chest injury
management [@sanddal2011analysis; @roy2017learning;
@o2013opportunities].

There is lacking evidence for the use of audit filters for improving trauma care 
[@evans2009audit; @stelfox2010quality]. Furthermore, the use of
audit filters has also been associated with high frequencies of false
positive OFI predictions, ranging from 24% to 80% [@sanddal2011analysis]; @roy2017learning], and some audit filters do not appear to correlate with
OFIs at all [@lewis2017delay].

## Aim

The aim of this study is to determine the performance of audit filters
in predicting OFIs and furthermore assess their relevance as a screening
tool for MMC.

# Methods

The research employed a single centre, registry-based, retrospective
cohort study with data from the Swedish Trauma Registry (SweTrau) and
the local trauma care quality database at Karolinska University Hospital
in Solna, to evaluate the performance of individual audit filters in
predicting OFIs.

## Study setting and population

Karolinska University Hospital is classified as a Trauma level I
hospital. Every year around 1,500 trauma patients receive treatment at
the hospital. If a patient case results in team activation it is added
to the Karolinska trauma registry. If no team activation occurs but the
patient has an Injury Severity Score (ISS) of less than 9
retrospectively, it is also included in the registry. The Karolinska
trauma registry reports to SweTrau. The registry includes data on vital
signs, times, injuries, and interventions as well as patient
demographics according to the European Consensus Statement, the Utstein
template [@ringdal2008utstein]

The Karolinska trauma registry also contains a local care quality
database including specific audit filters and OFIs that are determined
in MMC conferences through consensus decisions.

This process of determining OFIs unfolds in multiple stages,
characterised by escalating levels of scrutiny. Notably, instances of
mortality are directly referred to the multidisciplinary conference,
where, in addition to assessing OFIs, a decision is made regarding
whether the death was preventable or potentially preventable, a
classification also falling under the purview of OFIs.

From 2013 to 2017 there was an effort to identify adverse outcomes that
where unrelated to mortality, whereafter the review process underwent
subsequent refinement and formalisation during the study period. During
the initial period each trauma patient case underwent individual
assessments by a specialised trauma nurse to identify potential OFIs.
Only in 2017 was this procedure formalised and it became standard
practice to incorporate a preliminary individual evaluation by a
specialised trauma nurse upon data registration in the trauma registry
and the trauma quality database. The trauma quality data underwent
screening of audit filters. All cases falling within the criteria
delineated by these filters, along with those trauma patients flagged by
the nurse during the initial review for possible care failures, were
subjected to a secondary review by two specialised nurses. Subsequent
identification of a potential OFI during this second review prompted a
comprehensive evaluation of the relevant trauma patient's case.

## Participants

The study included all patient cases who underwent screening for OFIs.
Exclusion criteria were applied to those not screened for OFI and to
individuals under the age of 15, as their clinical management process
has notable distinctions compared to those applicable to the adult
population.

## Variables

### Outcome

The outcome variable in this study is an OFI, as established by the MMCs and furthermore treated as a dichotomous variable with "Yes -
At least one OFI identified" and "No - No OFI identified".

### Exposures

The 11 audit filters used at Karolinska University Hospital function as
the exposure variable. These filters can be seen in Table 1. Each audit
filter is labelled as either "original" or "manually created" depending
on how they were managed. Manually created audit filters were calculated
using trauma registry data instead of relying on the decision by the
specialised nurse as recorded in the quality database. The original
audit filters were instead used as recorded in the quality database,
since their criteria encompassed variables outside of what could be
retrieved from the existing data in the registry.

## Data sources/measurement

The data used in this retrospective cohort was retrieved from the
Karolinska University Hospital trauma registry and trauma care quality
database during the period from 2012 to 2022.

Patient data on vital signs, care processes and interventions, level of
care and time aspects was retrieved from SweTrau while both exposures
(audit filters) and outcome (OFI) were retrieved from the Karolinska
University Hospital trauma care quality database.

All data was anonymised to protect patient privacy.

## Bias

It is possible that consensus decisions from the MMC conferences are
flawed, leading to misclassification bias. Furthermore, the process of
flagging a patient case with an audit filter is done manually by a
specialised nurse as mentioned above. This means that there may be a
selection bias.

Health care professionals who attend MMCs and have also been a witness
to a specific patient case might not remember certain key elements that
may affect the decision-making around OFI. Patient case reviews in this
fashion are subject to recall bias. On the other hand, health care
professionals who have not witnessed that particular case may be over-
or underestimating the significance of an OFI.

## Study size

The study cohort encompasses all eligible patients treated at Karolinska
University Hospital from 2012 to 2022.

## Statistical methods

The study results were be generated by statistical analytics methods with the help of the statistical programming language, R 
[@R2022language].

Patient characteristics for patient cases with and without OFI were compared using Wilcoxon signed-rank test for continuous variables and a Chi-squared test for categorical values.

To evaluate the predictive performance of audit filters, the sensitivity and specificity of each individual audit filter in predicting OFI was calculated. Additionally, a receiver operating curve (ROC) for binary classification were constructed and the area under the curve (AUC) calculated in order to compare the performance of each audit filter 
[@robin2011proc].

Using a bootstrapping method with 1,000 iterations of the selected data, 95% confidence intervals for sensitivity, specificity and AUC were calculated. The significance of the AUC of each audit filter was then compared with that of coincidence (AUC = 0.5) and the minimum standard for diagnostic tests (AUC = 0.8) through an empirical bootstrap test with 1,000 iterations.

Instances of missing values for audit filters were treated as negative, as in "not flagged".

## Ethical considerations

In Sweden register-based studies do not require informed consent by patients in the registry. Approval by the Swedish Ethical Review Authority is required in order to execute a study in this manner. This authority comprises of medical, judicial and lay members appointed by the government. In order to uphold the principle of autonomy, measures are in place to safeguard patient dignity.

All patient data is anonymised according to General Data Protection Regulation (GDPR) standards [@gdpr2016EU]. The registry used provide data on various factors such as date and time of admission, patient characteristics e.g. age and gender. Theoretically, these attributes could make patient cases identifiable through external sources like news reports. Nonetheless, this means that such information must be leaked, which would only be possible through direct registry access. In order to minimise the risk of a potential data breach, the registry is password protected and access to the registry server is only possible while connected to a VPN with special credentials requiring two-factor authentication.

This was a retrospective cohort study and as such, no patients were put through any additional interventions and all patients were given care according to standard. Thus, patients were at no risk of any harm due the study.

This study will possibly give new insights to current quality improvement programs by assessing the advantage of audit filters. These new insights, if implemented, might further lead to a better outcome for trauma patients in terms of reduced mortality and morbidity.

This study was approved by Stockholm Research Ethics Review Board, approval number 2021-02541 and 2021-03531.

Table 1: All 11 audit filters used at Karolinska University hospital and included in this study. It shows which audit filters were created manually by registry data (manually created) and which audit filters were used as recorded in the trauma care quality database (original).

# Results

Figure 1 presents a flowchart showcasing the exclusion and inclusion of the patient cases from the trauma registry. After excluding patients under 15 years old and patients not screened for OFIs, a total of 8,309 individuals were included in the study. 7,797 out of 8,309 (93.8%) patient cases did not have an OFI and 512 out of 8,309 (6.2%) had an OFI.

## Missing Data

Table 2 shows the missing data for each audit filter. As mentioned in the method section, these values were interpreted as not flagged with an audit filter. The variables with the highest amount of missing data were “time to definitive treatment” (n = 5,990, 72.1%) and “time to first CT” (n = 1,012, 12.2%). The smallest amount of missing data was found in “SBP<90” (n = 13, 0.2%).

## Participant Characteristics

Table 3 presents the baseline characteristics of the study participants, comparing those with and without an assigned OFI. The variables presented in Table 3, except gender and age, were used in the calculation of audit filters.

Most trauma patients were men (n = 5,755, 69%). Patient characteristics varied depending on whether patients had an OFI or not. The mean age was significantly higher  
`r inline_text(tableOne, variable = pt_age_yrs, column = stat_2)` vs
`r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)`, `r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")` and
the ISS was significantly higher
`r inline_text(tableOne, variable = ISS, column = stat_2)` vs
`r inline_text(tableOne, variable = ISS, column = stat_1)`, `r inline_text(tableOne, variable = ISS, column = "p.value")`. The number
of intubations in the emergency department was significantly higher
`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_2)`
vs
`r inline_text(tableOne, variable = ed_intubated, level = "No", column = stat_1)`, `r inline_text(tableOne, variable = ed_intubated, level = "No", column = "p.value")`.
Patients with OFI had longer times to definitive treatment from hospital
arrival compared to patients without OFI
`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2)`
vs
`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_1)`, `r inline_text(tableOne, variable = dt_ed_emerg_proc, column = "p.value")`.
The time to the first CT was also longer in patients with OFIs compared to patients without
`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_2)` vs
`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_1)`, `r inline_text(tableOne, variable = dt_ed_first_ct, column = "p.value")`.


The number of OFIs were highest in patients admitted to intensive care unit `r inline_text(tableOne, variable = host_care_level, level = "Intensive care unit", column = stat_2)` and surgical ward `r inline_text(tableOne, variable = host_care_level, level = "Surgical ward", column = stat_2)`.

Gender `r inline_text(tableOne, variable = Gender, column = "p.value"`, the mean systolic blood pressure `r inline_text(tableOne, variable = ed_sbp_value, column = "p.value"`, death at 30 days `r inline_text(tableOne, variable = res_survival, column = "p.value"` and trauma team activation `r inline_text(tableOne, variable = Tr_Nivå, column = "p.value"` was not significantly correlated with OFI.

furthermore there was a statistical significant correlation between GCS at ED and OFI (p < 0.001)

## Individual Audit Filters

### Performance of Individual Audit Filters

The performance of each audit filter was determined through a calculation of sensitivity and specificity. This is demonstrated in Table 4.

Varying numbers of patient cases were flagged with each audit filter. The audit filter flagged most frequently was “>30 min until CT”. The least flagged audit filter was “GCS<9 and not intubated”.

The audit filter with the highest sensitivity was “>30 min until CT” (59.2%, 95% CI [55.2-63.8]). The audit filter with the lowest sensitivity was “GCS<9 and not intubated” (1.2%, 95% CI [0.2-2]).
The audit filter with the highest specificity was “GCS<9 and not intubated” (98.7%, 95% CI [98.5-98.9]). The audit filter with the lowest specificity was “>30 min until CT” (53.8%, 95% CI [52.7-54.8]).

Generally, the confidence intervals varied between sensitivity and specificity, with sensitivity having a greater interval. 

Table 5 summarises the performance of each individual audit filter in identifying OFIs in trauma care by AUC. Four of the 11 audit filters,"SBP<90", “dead at 30 days”, “GCS<9 and not intubated” and “CPR and thoracotomy” showed a performance in predicting an OFI equal to that of coincidence (AUC = 0.5). ">60 min until first intervention" had the highest AUC (0.63, 95% CI [0.61-0.65]). All Audit filters showed significantly lower AUC compared to the minimum standard for diagnostic tests (p < 0.001). 

Table 2: Comparison of missing values for each audit filter. Missing values is shown as a number
and percentage of the total number of patient cases included in the study (n = 8309).\
`r tableTwo` Table 3: Patient characteristics of individuals with and
without an OFI. Missing values are shown only when they exist.
`r tableOne` Table 4: Comparison of performance for each audit filter in
predicting OFIs. Performance is defined by sensitivity (%) and
specificity (%). A 95% confidence interval is shown together with each
performance indicator. The number of cases flagged by each filter is
displayed.\
`r tableThree` Table 5: Comparison of performance for each audit filter
in predicting OFIs. Performance is defined by the area under the
receiver operating characteristics curve. P-values show the significance
of AUC compared to coincidence (AUC = 0.5) and to the minimum standard for diagnostic tests (AUC = 0.8) through an empirical bootstrap test. A 95%
confidence interval is used. 
`r tableFour`

```{r fig1, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE, error=FALSE}
knitr::include_graphics("ofi-flowchart.pdf")
```

Figure 1: Flowchart of the patient selection process. Shows the patients
excluded and included in this study.

# Discussion

The aim of this study was to evaluate the performance of individual audit filters in predicting OFIs and to assess their suitability as a screening tool for morbidity and mortality conferences. This study showed that audit filters do not perform well in predicting OFIs.

## Interpretation of Results

The observed twofold prevalence of male trauma patients compared to females is consistent with previous findings both nationally
[@SweTrau2022Arsrapport] and globally [@roth2018global]. This is believed to lie in the tendency toward high-risk behaviours in males.

A previous study conducted at Karolinska University Hospital by Albaaj et al. [@albaaj2023patient] sought to identify patient factors associated with OFIs and found that a higher ISS, longer times to CT, longer times to first intervention and higher levels of care was correlated with OFIs. These findings correspond with our analysis of patient characteristics.

Higher ISS and higher levels of care in the patients with OFIs as opposed to patients without OFIs could be explained by the greater complexity of such trauma cases. Greater complexity requires more interventions which could allow for more mistakes [@albaaj2023patient].

Notably, GCS at ED showed a correlation with OFI eve though the median GCS and its IQR was identical for individuals with OFI and without. Upon further analysis this could be explained by the difference in distribution outside the IQR and due to a higher relative amount of missing data.   

Audit filters generally exhibited a high specificity relative to their sensitivity which may be attributed to the outcome being rather uncommon, whilst every individual audit filter misses the majority of OFIs. Theoretically, combining all audit filters together as one audit filter would result in the sensitivity increasing, and simultaneously the specificity decreasing.

An extensive review by Evans et al. [@evans2009audit] sought to review the evidence for audit filters as a tool to improve care. A total of 741 studies were screened by title. Of these studies, 42 abstracts were appraised and five complete articles were examined in detail. None of the five studies met the inclusion criteria established. Two of the studies by Chadbunchachai et al. 
[@chadbunchachai2001comparative; @chadbunchachai2003study], which were
examined in detail by Evans et. al were mentioned in WHO:s guidelines
for quality improvement programmes [@world2009guidelines].They were excluded from the Evans et al. review due to no clear pre‐ or post‐intervention data points or a clearly defined time when the respective interventions were implemented.

In 2008, Willis et al. [@willis2008interpreting] assessed 14 audit
filters, finding only three to be associated with an increased risk of
poor outcomes. One audit filter, "GCS\<9 and not intubated", also used
in this study, similarly showed no significant ability to predict poor outcome.

Audit filters demonstrating a performance equal to coincidence are deemed unsuitable as screening tools for OFIs. On the contrary, audit filters demonstrating a performance superior to that of coincidence could still be used as audit filters. However, it is important to note that despite these audit filters showing correlation with OFI, it does not mean they are useful as a screening tool. That would not have been the case with a diagnostics test where the AUC should be at least 0.8 for it to be considered acceptable  [@nahm2022receiver].

A noteworthy challenge with audit filters is the risk of over-reliance on them. While they may help in determining the existence of OFI, they could overshadow other factors that may contribute to that analysis.

Audit filters are static in their criteria, contrary to health care in general. This means that audit filters may lose their utility if new guidelines are introduced. For example, liver and spleen injuries did not perform at all in predicting OFIs.

A further problem the static nature of audit filters is that they can fulfil their purpose and subsequently lose their relevance through a negative feedback loop. For example, a problem in health care occurs and in response, an audit filter is designed to tackle the problem. At first this will theoretically result in a good predictability for OFI. However, when such OFIs are taken into consideration the performance of the audit filter will gradually weaken.

Some audit filter criteria are more rigid than others. A criterion too rigid will fail to recognise the majority of OFIs while too inclusive criteria might flag several patients, who do not necessarily have an OFI. This in turn leads to further issues regarding the resources and time needed for making an OFI decision. A very inclusive criterion will cause many patient cases to be flagged and will lead to many patient cases being discussed at MMCs.

Suppose we were to replace audit filters used by Karolinska University Hospital with other audit filters. We would have to review them over time and simultaneously risk the possibility that they might not perform.

Another general challenge with audit filters is the unintended bias that comes with devising audit filters. They might disproportionally affect certain groups of people. Moreover, the risk of confounders is present when establishing such criteria.

Audit filters currently in use at Karolinska University Hospital focus primarily on aspects of trauma care mostly relevant to medical doctors. It might be beneficial to consider the interdisciplinary nature of trauma care by using criteria related to morbidity for example.

### Delay

Both audit filters measuring delay: "\>60 until first intervention" and
"\>30min until first CT", showed a correlation with OFI. This is
comparable to what Teixeira et al. [@teixeira2007preventable]
demonstrated, with delay correlating to preventable death. Furthermore,
Ghorbani et al. [@ghorbani2018analysis] identified delay in treatment,
delay to CT, and clinical judgment errors as common errors in trauma
care. A more recent study in Japan by
Yamamoto et. al [@yamamoto2023immediate] showed a significant reduction in mortality when patients received a whole-body CT within 10 minutes of arrival at the ED. It is important to note that OFI in this study was not only a reflection of mortality but a broader concept encompassing different aspects of care quality. If anything, it highlights the difference between hospitals, which audit filters are in use, and how they are being assessed.

The audit filters “Dead at 30 days” and “CPR and thoracotomy” showed no difference from coincidence (AUC = 0.5) in predicting OFI, potentially due to severe traumatic injuries inevitably resulting in death within 30 days of admission to the hospital, and presenting in a smaller time frame for mistakes to be made.

Albaaj et al. [@albaaj2023patient] examined patient factors associated with OFIs, identifying delays of 30-60 minutes until the first CT, higher levels of care and higher ISS as linked to OFIs. Similarly, they found no clear association between 30-day survival and OFIs. This could partly explain the correlation between “ISS > 15 and no team activation” and OFI and similarly regarding “ISS > 15 and not in intensive care unit”.

Sanddal et al. [@sanddal2011analysis] highlighted that OFIs predominantly exist in the ED and relate to airway management, fluid resuscitation, and chest injury management. The only audit filter directly associated with airway management was “GCS under 9 and not intubated” which performed comparably with coincidence (AUC = 0.5) in predicting OFIs. This could suggest that an alternative audit filter related to airway management might potentially perform differently.

The purpose of audit filters is to elevate the quality of care. Audit filters should therefore reflect the current trauma care quality gap. However, once quality is improved, these audit filters lose their purpose. This could suggest that rather than the audit filters performing poorly, they have actually fulfilled their efficacy, and consequently their performance weakens. It is therefore crucial to continuously update the audit filters.


## Generalizability

Given the variability in screening for OFI, it is likely that results will also vary depending on the composition of an OFI and the specific audit filters used at a given trauma centre. Likewise, audit filters are used to fill quality gaps and due to the unique quality gaps present at individual trauma centres, there will be no standard set of universal audit filters. The audit filters used in this study might not perform similarly in other trauma systems. 

Audit filters and quality improvement programmes are widely used and the findings in this study suggest that audit filters generally perform poorly in predicting OFI. The audit filters with performance exceeding coincidence (AUC = 0.5) might be a beneficial addition to an already existing quality improvement programme or could be used to fine-tune audit filters already in use to plug hospitals’ quality gaps. Due to the study being conducted on a single, trauma level I centre in Stockholm it is unrealistic to assume applicability of these findings at other hospitals in other countries facing different challenges in trauma care. It has already been suggested that the efficacy of audit filters will vary depending on the country in question. Low and middle income countries seem to benefit from certain audit filters that high income countries do not and vice versa [@berg2022perceived].

## Strengths

As stated in the method section, some audit filters were manually created from registry data instead of relying on the selection process by the specialised nurse. This, in order to minimise misclassification bias.

Furthermore, there has been very limited high quality evidence supporting the efficacy of audit filters  in improving trauma care [@evans2009audit; @stelfox2010quality] and predicting OFI [@sanddal2011analysis; @roy2017learning]. This study shows comparable results.

Previous studies focus primarily on audit filters’ ability to reduce mortality [@chadbunchachai2001comparative; @chadbunchachai2003study]. In this study the morbidity of patient cases is also taken into account using OFIs as an umbrella term that includes factors ranging from bad documentation to preventable death.

From 2017 all trauma patients admitted to Karolinska University Hospital were included in the registry reducing the risk of bias in inclusion of patient cases.


## Limitations

Opportunity for improvement, while dichotomously defined, includes a diverse set of outcomes. This makes it difficult to create audit filters that perform well in predicting OFIs.
The current screening system for OFIs might also introduce bias since the audit filters would favour the identification of some, but not all, errors.

As stated in the method section, the performance of audit filters was calculated by reviewing every patient case with or without an OFI and simultaneously flagging or not flagging with the audit filter. Consequently, even if an audit filter demonstrated a true positive correlation with an OFI, instances where the same patient case had multiple audit filters introduce a degree of uncertainty regarding which specific audit filter accurately predicted the outcome.

The number of missing values varied greatly among the audit filters. As stated above, missing values from each audit filter were coded as “not flagged” by an audit filter. This means that some instances where patient cases were not flagged, might actually meet the audit filter criteria, but for some reason not documented. This reduces the validity of some audit filters that featured high amounts of missing data.

## Future Studies

Given the expanding field of artificial intelligence and machine learning, newer studies should focus on using these tools to screen for patient cases at risk of an OFI. This will reduce misclassification and ensure constant and continued good performance due to an algorithm’s ability to learn from the current context of trauma care. This approach is new but not unexplored. A preprint is currently available by Attergrim et al. 
[@attergrim2023predicting] exploring the idea of machine learning models as a screening tool for potential OFIs. They found that machine learning models outperformed the audit filters currently used at Karolinska University Hospital in predicting OFIs.

If the use of an audit filter is still considered, it is important to ensure its performance. Future studies could look at subgroups of OFIs, and which audit filter responds to that subgroup. Some audit filters might perform well in predicting a subgroup of OFIs but fail to predict others.


## Health Equity

It could be of interest to study the occurrence of OFIs among marginalised groups of people and women in order to assess whether some groups tend to excluded from the screening method or are at a higher risk of being falsely labelled with or without OFIs.

There is a growing body of research [@schoenfeld2013patient; @bedri2017national] suggesting that certain populations have poorer health and treatment outcomes due to implicit biases that may affect screening and treatment by health care professionals, sometimes leading to unequal distributions of mortality and morbidity among minority groups. Reviewing OFIs in adult trauma patients from this perspective could help uncover and address health inequities in this regard.

Since there is a gender disparity between the number of trauma patients, it could be worthwhile to look further into whether gender disparities in OFIs could reveal gender discrimination in the screening and labelling of OFIs.


# Conclusion

Audit filters are widely used as a tool in quality improvement programmes. Strong evidence supporting their efficacy in quality improvement is scarce, but this seems to depend on where and when they are used, in what context and for what purpose. In this study, no audit filter seemed to show good performance in predicting OFIs with AUC ranging from 0.5 to 0.63. These results highlight the importance of continually evaluating the relevance of certain audit filters. Audit filters should reflect the current gap in trauma care quality. This process takes time and resources and using an alternative to audit filters is recommended.

# Acknowledgements

I wish to thank my supervisor Martin Gerdin Wärnberg and my co-supervisors Jonatan Attergrim and Kelvin Szolnoky for their support and guidance throughout the entirety of this study.

# References

::: {#refs}
:::

