---
title: Performance of individual audit filters in predicting opportunities for improvement
  in adult trauma patients
author: "Josef Al-Khalili"
subtitle: A registry based cohort study
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
csl: bmcemerg.csl
editor_options:
  markdown:
    wrap: 72
bibliography: bibliography.bib
always_allow_html: yes
---
```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
source("main.R")
library(rofi)
library(dplyr)
library(ggplot2)
library(pROC)
library(gtsummary)
library(gt)
library(boot)
library(webshot2)
library(knitr)
```

# Abbreviations

OFI – Opportunities for Improvement 

CI – Confidence Interval 

GCS – Glascow Coma Scale 

ISS – Injury Severity Score 

ED – Emergency Department 

AUC – Area Under Curve 

ROC – Receiver Operating Characteristics 

CPR – Cardiopulmonary Resuscitation 

SBP – Systolic Blood Pressure 

WHO – World Health Organisation 

YLL – Years Life Lost

DALY – Disability Adjusted Life Years 

PTSD – Post Traumatic Stress Disorder

<!-- Add abstract -->

# Introduction

Trauma is defined as a physical injury of external origin together with
the associated reaction of the body. This phenomenon accounts for 9% of
global deaths annually [@world2014injuries]. Trauma affects low- and
middle-income countries at a higher rate. In addition, young people are
at higher risk of trauma compared to the general population. It is the
leading cause of death in people between the ages of 10 and 49
<!-- If you want to combine citations you should write
[@vos2020global; @roth2018global] -->
[@vos2020global];[@roth2018global]. The most common type of trauma for
people of all ages is road traffic injuries
[@vos2020global];[@roth2018global]. In Sweden, road traffic injuries
account for about 40% of injuries and falls for 40%
[@SweTrau2022Arsrapport].

Trauma is broadly divided into two groups according to the underlying
causal mechanism: penetrating and blunt trauma. Common examples of
penetrating trauma are gunshot wounds and knife cuts. Blunt trauma, on
the other hand, is caused by injuries such as road traffic accidents and
falls. In general, the two types of trauma can also be characterized by
the size of the force causing the injury and the area of impact. Blunt
trauma is caused by a high force impact over a larger area. In contrast,
penetrating trauma is caused by a force applied to a smaller area,
causing greater pressure so that it is penetrating and often results in
deeper lacerations to tissues and organs. However, blunt trauma with
sufficient force can be penetrating in nature [@hexdall2011Mechanism].
Trauma varies widely and has both long term and short term consequences on an individual, a societal level and economical level.  

## Impact
<!-- Consider merging this section with the paragraphs above, as they
are all about the characteristics and impact of trauma -->
The effects of a trauma depend greatly on the type of trauma, where it
happened and whom it affected.

Trauma is responsible for 11.9% of years of life lost (YLL) worldwide
[@roth2018global]. In addition, trauma is associated with a higher
propensity for depression, post-traumatic stress disorder (PTSD) and
negative effects on cognitive function, etc.
[@north1999psychological];[@muscatelli2017prevalence].

<!-- Do you mean the same thing by "trauma" and "traumatic injuries"? 
If so, I suggest that you stick to just "trauma"-->
It is estimated that up to 11% of trauma patients have PTSD 3 years
after the traumatic event [@visser2017course]. In addition, people who
witness traumatic injuries are also at a higher risk of developing
PTSD. Traumatic injuries are associated with 36% not returning to work within
1 year [@david2022measuring].

<!-- By "injuries", do you mean the same thing as "trauma"? Be consistent in
how you use words, stick to one -->
However, injuries are not only a burden on the person affected, but also
on society as a whole [@newnam2014impacts]. One aspect of such societal
impact is the economic cost. Attempts have been made to estimate the
global cost of injuries, which is a complicated task due to their
variability. However, it is estimated that the cost of road traffic
injuries alone accounts for 2% of gross domestic product (GDP) in
high-income countries [@world2014injuries].

## Trauma Care Systems

<!-- Trauma care is not only provided by specialised trauma centres.
A lot of trauma care is provided in general hospitals, so I suggest 
that you remove that part -->
Trauma care is a multidisciplinary, complex and time-critical healthcare
service provided by specialised trauma centres. A variety of different
specialties are involved in trauma care, including surgeons,
orthopedists, anesthesiologists, emergency physicians and nurses. Trauma
care is divided into three components: pre-hospital, in-hospital and
post-hospital. The prehospital component constitutes first responders, basic life support and advanced life support.[@world2005prehospital]. The in hospital component is defined as the instance a trauma patient is admitted to a health care facility, often a trauma center. The post-hospital component is about the rehabilitation and recovery of treated trauma patients.  

Trauma centres have a significantly lower mortality rate compared to non-trauma centres, and
the maturity of a trauma centre correlates with lower mortality,
demonstrating the importance of high-quality, specialised trauma care
[@alharbi2021effectiveness];[@mann1999systematic]. A similar study in
Sweden [@candefjord2020mortality] showed that 30-day mortality is 41%
lower in a mature trauma centre compared to a non-trauma centre.
<!-- I don't think it was mortality that was 41% lower, but the adjusted 
risk of death. This means that if you arrive to a trauma centre your risk of 
death is 41% lower, all other things equal. But the actual mortality is much lower -->
However, even established trauma care systems are prone to error due to its
complexity and susceptibility to delays
[@nikouline2021errors];[@vioque2014classifying], which in turn can lead
to complications and death. Studies between 1990 and 2014 have shown a
pooled preventable death rate of 20%, with more recent studies showing
less [@kwon2014pooled]. A recent study in a Swedish Level I trauma
centre found a preventable death rate of 4% [@ghorbani2018analysis].

## Quality Improvement

To minimize errors in trauma care, quality improvement (QI) programs
have been established by the World Health Organization (WHO) and the International
Association for Trauma Surgery and Intensive Care (IATSIC)
[@world2009guidelines]. These programs form the core of current trauma
care systems and their development around the world.

QI programs consist of a series of recommended techniques aimed at
facilitating discussion and analysis of the trauma process in order to
take corrective action for future success. The following techniques are
used: A multidisciplinary mortality and morbidity conference (MMC), a
preventable death review panel, and the use of audit
filters[@world2009guidelines].

### Morbidity and Mortality Conferences and Preventable Death Review Panels

The MMC is a central QI technique. It is a proven framework with the
critical components of anonymity, focusing on specific adverse events,
recognising flawed approaches, critical analysis, seeking change and
implementing that change [@gregor2016morbidity].

MMCs are regular meetings at which specific patient cases are presented
for discussion. The conference participants are healthcare professionals
involved in trauma care and its processes, but not necessarily the
specific patient case. The selection of patient cases varies, but common
themes for discussion are unexpected mortality, unexpected morbidity and
errors [@orlander2003morbidity].

It is recommended that they last about 40 minutes, but this varies from
hospital to hospital [@world2009guidelines].

Five themes have been proposed that broadly define QI-centred MMC: (1)
clear definition of the role of the MMC (2), involvement of
stakeholders, (3) recognising and selecting appropriate patient cases
for presentation, (4) structuring<!-- Do you mean structured?--> focused discussions, and (5)
formulating recommendations and assigning follow-up
actions[@giesbrecht2016morbidity];[@churchill2020quality].

When implemented effectively, MMCs serve as an important tool for error
proofing without serving solely as a negative for individual care
providers, but rather as an incentive for reporting. These conferences
promote a more comprehensive, systems-based approach that emphasizes
task analysis, teamwork and quality improvement[@lazzara2022morbidity].
The introduction of MMC has been shown to increase morbidity and
mortality reporting[@kong2016analysis].

<!-- I suggest you remove this part about the avoidable death panel, as it
is specific to KUH -->
Following MMC, there is an avoidable death review panel whose role is to
determine whether a patient's death could have been avoided
[@ghorbani2014clinical]. 

### Audit filters and Opportunities for improvement

Audit filters also known as quality indicators are specific established
criteria involved in trauma processes and care. They are used to detect
deviations from standardized care in order to further analyse the cause
behind complications in trauma patients. An audit filter can be death
and the placement of two large bore intravenous lines within 15 minutes
from arrival to a healthcare facility [@world2009guidelines]. The
purpose of audit filters is to improve the quality of care.

Earlier studies showed a reduction in trauma related mortality after the
implementation of audit filters [@chadbunchachai2001comparative].
However, newer studies seem to differ. In 2009 there was an extensive
review made that sought to determine if audit filters could be used in
improving processes of trauma care however none of the studies met
established inclusion criteria [@evans2009audit]. The use of audit
filters have also been associated with high frequencies of false
positives, ranging from 24% to 80%.[@sanddal2011analysis];
[@roy2017learning].

OFIs are the endpoint and aim of MMC. Following the review of individual
patient cases there is a consensus decision made regarding the existence
of an OFI. It has been shown that such review process is associated with
high-quality trauma care [@santana2014development]. OFIs are typically
associated with failures in initial care [@vioque2014classifying],
specifically in airway management, fluid resuscitation, haemorrhage
control and chest injury management [@sanddal2011analysis];
[@roy2017learning]; [@o2013opportunities]. Some audit filters seem to
not correlate with OFIs at all [@lewis2017delay]

## Aim

The aim of the study is to determine the performance of audit filters in predicting OFIs.

# Methods

We conducted a single center, registry-based, retrospective cohort study with data from the
trauma registry and trauma care quality database at the Karolinska
University Hospital in Solna in order to evaluate the performance of
individual audit filters in predicting OFIs.

## Study setting and population

Karolinska University Hospital is classified as a Trauma level I
hospital. Every year around 1500 trauma patients receive treatment at the
hospital. If a patient case results in team activation it is added to
the Karolinska trauma registry. If no team activation occurred but the
patient had ISS>9 retrospectively it is also included in the registry.
The Karolinska trauma registry reports to the Swedish Trauma registry
(SweTrau). The registry includes data on vital signs, times, injuries,
and interventions as well as patient demographics according to the
European consensus statement, the Utstein template [@ringdal2008utstein]

The Karolinska trauma registry also contains a local care quality database
including specific audit filters and OFIs that are determined in MMC
conferences through consensus decisions.

This process of determining OFIs unfolds in multiple stages,
characterized by escalating levels of scrutiny. Notably, instances of
mortality are directly referred to the multidisciplinary conference,
where, in addition to assessing OFIs, a decision is made regarding
whether the death was preventable or potentially preventable, a
classification also falling under the purview of OFIs.

From 2013 to 2017 there was an effort put forward in identifying adverse
outcomes that where unrelated to mortality, whereafter the review process
underwent subsequent refinement and formalization during the study
period. During the initial period each trauma patient case underwent
individual assessments by a specialized trauma nurse in order to
identify potential OFIs. It was, however not until 2017 in which this
procedure was formalized and it became standardized to incorporate a
preliminary individual evaluation by a specialized trauma nurse upon
data registration in the trauma registry and the trauma quality
database. The trauma quality data underwent screening of audit filters.
All cases falling within the criteria delineated by these filters, along
with those trauma patients flagged by the nurse during the initial
review for possible care failures, were subjected to a secondary review
by two specialized nurses. Subsequent identification of a potential OFI
during this second review prompted a comprehensive evaluation of the
respective trauma patient's case.

## Participants

In this study, we included all who underwent screening for OFIs.
Exclusion criteria were applied to individuals under the age of 15, as
their clinical management process has notable distinctions compared to
those applicable to the adult population.

## Variables

### Outcome

The outcome variable in this study is an OFI, as established by the MMC
conference and furthermore treated as a dichotomous
variable with "Yes - At least one OFI identified" and "No - No OFI
identified".

### Exposures

The exposure variable is the 11 audit filters used at Karolinska
University Hospital.

## Data sources/measurement

The data used in this retrospective cohort was retrieved from the trauma registry and the
trauma care quality database during the period spanning from 2012 to
2022.

Patient data on vital signs, care processes and interventions, level of care and time aspects etc. was retrieved from the national trauma registry while both exposures (audit filters) and outcome (OFI) was retrieved from the local trauma care quality database. 

All data were anonymized to protect patient privacy. 

## Bias

There is a possibility that consensus decisions from the MMC conferences
are flawed which causes misclassification bias. Furthermore, the process
of flagging a patient case with an audit filter is done manually by a
specialised nurse as mentioned above. This means that there may be a
selection bias.

There is a level of recall bias in these sort of studies. Care professionals who attend the MMC and has also been a witness
to the specific patient case in question might not remember certain key
elements that might affect the decision-making around OFI. 

## Study size

The study cohort encompasses all eligible patients treated at the
hospital from 2012 to 2022. 

## Statistical methods

The study results are generated by statistical analytics methods with
the help of the statistical programming language, R. [@R2022language]

To evaluate the predictive performance of audit filters, the sensitivity and specificity of each individual 
audit filter in predicting OFI will be calculated. Furthermore, a receiver 
operating curve (ROC) for binary classification were constructed and the area under the
curve (AUC) calculated in order to compare the performance of each audit
filter [@robin2011proc]

<!-- Be careful about using correct tense. You mix future and past tense, but 
when you're done you should use past tense consistently -->
Using a bootstrapping method with 1000 iterations of the selected data,
 95% confidence intervals for sensitivity, specificity and AUC were
calculated. The significance of the AUC of each audit filter will then be compared with that of
coincidence (AUC = 0.5), through a one sample T-test. 

<!-- Consider introducing the filters earlier, perhaps under "Exposure"-->
Table 1 shows a list of the audit filters currently in use at Karolinska university hospital. 
Each audit filter is also labelled with either original or manually created. 
following audit filters: "SBP<90", "GCS < 9 and not
intubated", ">30 min until first CT", ">60 min until first
intervention", "ISS>15 and not in intensive care unit", "ISS>15 and
no team activation", "death in 30 days" were manually generated from registry data to 
minimize selection bias during the specialized nurse's manual flagging of audit filters. 
It is noteworthy that the remaining audit filters cannot be derived from the existing data in the registry.
<!-- The distinction between the "original" and "manually generated" filters need to be 
clarified. You mean that some were assessed as they were recorded in the quality database whereas 
others where calculated from the trauma registry data?-->

Instances of missing data for audit filters were treated as indicative of the absence of the respective audit filter.
<!-- Do you mean that they were treated as negative, i.e. not flagging a specific patient? -->

## Ethical considerations 

All patient data is anonymized. However there is data on
various factors like date, time of admission, etc. and different patient characteristics such as age, gender, etc. Theoretically, these attributes could make patient cases identifiable through external sources like news reports. However this means that such information has to be leaked which would only be possible through direct registry access. In order to minimize the risk of a potential data breach, access to the registry is only possible while connected to a VPN with a special credentials.

<!-- Please review the KI guidelines and check that what you have here is adequate. 
If not, consider discussing the ethical considerations using the different principles
of medical research ethics -->

The study was approved by
Stockholm Research Ethics Review Board, approval 107<!-- Where does this 107 come from?--> number 2021-02541
and 2021-03531. 

# Results 

Figure 1 presents a flowchart showcasing the exclusion and inclusion of the patient cases from the trauma registry. After excluding patients under 15 years old and patients not screened
for OFI, a total of 8309 individuals were included in the study. 7797 out of 8309 (X%)<!-- Consider adding % here--> patient cases did not have an OFI and 512 out of 8309 <!-- Same here -->patient cases had an OFI.

## Participant Characteristics

Table 2 presents the baseline characteristics of the study participants, comparing those with and those without an assigned OFI. The variables presented in Table 2 exhibited a direct correlation with the manually created audit filters. <!-- Not sure I understand this last sentence. Do you mean that Table 2 shows the manually created filters?-->

<!-- Add numbers as well-->
69% out of all patients included in the study where men and 31% where women. 
Among the patients with an OFI compared to the  patients
without, the mean age was slightly higher `r inline_text(tableOne, variable = pt_age_yrs, column = stat_2)` vs `r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)` and
the ISS was higher `r inline_text(tableOne, variable = ISS, column = stat_2)` vs `r inline_text(tableOne, variable = ISS, column = stat_1)`. The number of intubations in the emergency department was also
higher `r inline_text(tableOne, variable = ed_intubated, column = stat_2)` vs `r inline_text(tableOne, variable = ed_intubated, column = stat_1)`. Patients with OFI had longer times to definitive
treatment from hospital arrival compared to patients without OFI `r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2)` vs `r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_1)`. Furthermore the time to first CT was also longer
in the group with OFI compared to the group without: `inline_text(tableOne, variable = dt_ed_first_ct, column = stat_2)` vs `r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_1)`. 

The mean systolic blood pressure in the OFI group did not differ from
the systolic blood pressure in the no OFI group.

Table 2 shows the missing data for each audit filter. As mentioned in the method section, these values where interpreted as not flagged with an audit filter.
The variables with the highest amount of missing data was "time to
definitive treatment" [n=5990] and "time to first CT" [n=1012].

<!-- Note that you haven't numbered sections before here -->
## 2. Individual Audit Filters

### 2.1. Performance of Individual Audit Filters

The performance of each audit filter was determined through a calculation of sensitivty and specificity. 
Table 3 summarizes the performance of each individual audit filter in
identifying opportunities for improvement in trauma care by specificity
and sensitivity. <!-- This is a bit repetitive, consider merging the two sentences -->

There where varying numbers of patient cases flagged with each audit
filter. The audit filter that was flagged most frequently was ">30 min
until CT". The least flagged audit filter was "GCS<9 and not
intubated".

<!-- I recommend that you present sensitivity and specificity in the 
same order always, otherwise you will confuse your readers. I would suggest
sensitivity first and then specificity -->
The audit filter with the highest specificity was "GCS<9 and not
intubated" (98.7% CI: 98.5-98.9) was also the audit filter with the
lowest sensitivity (1.2% CI: 0.1-2).

The audit filter with the highest sensitivity was ">30min until CT"
(59.2% Ci: 55.3-63.3). The same audit filter had the lowest specificity
(53.8% CI: 52.7-55.2).

The confidence intervals varied between sensitivity and specificity with
sensitivity having greater span.
<!-- Do you mean for all filters? Do you mean that sensitivity estimates, 
in general, where associated with larger uncertainty?-->

Table 5 summarizes the performance of each individual
audit filter in identifying OFIs in trauma care by AUC.
Three of the ten audit filters, "Dead at 30 days", "GCS<9 and not
intubated" and "CPR and thoracotomy" showed a performance in predicting
an OFI as good as flipping a coin. >60 min until first intervention
had the highest AUC. 

## Missing Data 
<!-- Missing data is often presented earlier in results, maybe even the
first paragraph-->
The audit filter that displayed the highest amount of missing data was
">60 min until first intervention" (n = 5990, 72.1%). The least amount
of missing data was found in "SBP<90" 

Table 2: Patient characteristics of individuals with and without an OFI. Missing values is shown only were there is any. 
`r tableOne`
Table 3: Comparison of missing values for each audit filter as a number and as a percentage of the total amount of instances.  
`r tableTwo`
Table 4: Comparison of performance for each audit filter in predicting OFI. Performance is defined by sensitivity (%) and specificity (%). A 95% confidence interval is shown together with each performance indicator. The number of cases flagged by each filter is displayed.     
`r tableThree`
Table 5: Comparison of performance for each audit filter in predicting OFI. Performance is defined by the area under the receiver operating curve. P-value show significance of AUC compared to 0.5 through t-test. A 95% confidence interval is used.
`r tableFour`
```{r fig1, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE, error=FALSE}
knitr::include_graphics("ofi-flowchart.pdf")
```
Figure 1: Flowchart of the patient selection process. Shows the patients excluded and included in this study. 

# Discussion 

## Interpretation of Results 

The aim of the study was to evaluate the performance of individual audit filters in predicting OFI to assess their suitability as a screening tool for morbidity and mortality conferences.
<!-- Before you start discussing how demographics differed between groups I suggest that you summarise your key findings. What were the most important findings? That all filters performed poorly? -->

The observed twofold prevalence of male trauma patients compared to females is consistent with previous findings both nationally [@SweTrau2022Arsrapport] and globally [@roth2018global]. It is thought that the reason for this lies in the higher-risk behaviors by males.

The higher ISS in the OFI group as opposed to the group without OFI could be explained by the greater complexity of trauma cases with increasing ISS. Greater complexity allows for more mistakes. These findings stay true with a previous study conducted at Karolinska University Hospital by Albaaj et. al [@albaaj2023patient]. Albaaj et. al sought to identify patient factors associated with OFI and found that a higher ISS was significantly correlated with OFI.

Audit filters generally exhibited a high
specificity relative to their sensitivity which may be attributed to the outcome being rather uncommon whilst every individual audit filter miss the
majority of OFI. Theoretically, adding all audit filters together as one audit filter would
result in the sensitivity increasing.

An extensive review by Evans et. al [@evans2009audit] sought to review the evidence for audit filters as a tool to improve care. They screened for a total of 741 studies by title. Of these studies, 42 abstracts were appraised and five complete articles were examined in detail. None of the five studies met the inclusion criteria established. two of the studies which where examined in detail by Chadbunchachai et. al [@chadbunchachai2001comparative];[@chadbunchachai2003study] where also two of the studies mentioned in WHO:s guidelines for quality improvement programmes [@world2009guidelines]. They were excluded due to no clear pre‐ or post‐intervention data points or clearly defined a point in time when the respective interventions were implemented.

In 2008, Willis et. al [@willis2008interpreting] assessed 14 audit filters, finding only three to be associated with an increased risk of poor outcomes. One audit filter, "GCS<9 and not intubated", also used in this study showed no significant ability to predict poor outcome.

The audit filters demonstrating a performance equal to coincidence are deemed unsuitable as screening tools for OFI.
The audit filters demonstrating a performance superior to that of coincidence could still be used as audit filters. However, it is important to note that even though these audit filters proved meaningful does not mean they are acceptable as a screening tool as that would not have been the case with a diagnostics test where the AUC should at least be 0.8 for it to be considered acceptable [@nahm2022receiver]

<!-- There's considerable scope to expand on why you think that they perform so poorly, along the lines of our discussions during the meeting yesterday. Is it because they've played out there role? Is it because the care has improved to the point that OFIs are no longer related to these filters? What can happen if we replace them with new filters? Is that a feasible and reasonable approach?-->

### Delay

Both audit filters measuring delay: ">60 until first intervention" and ">30min until first CT", showed relatively good performance in predicting OFI<!-- This is an overstatement if we're talking an AUC in the realm of 0.6 -->. This is comparable to what Teixeira et. al [@teixeira2007preventable] demonstrated, with delay being correlated with preventable death. Furthermore Ghorbani et al. [@ghorbani2018analysis]  identified delay in treatment, delay to CT, and clinical judgment errors as common errors in trauma care. This aligns with our findings. A more recent study in Japan by Yamamoto et. al [@yamamoto2023immediate] showed a significant reduction in mortality when patients received a whole body CT within 10 minutes of arrival to the ED. A german study by Bieler et. al [@bieler2020evaluation] evaluated quality indicators through a survey partaken by different medical professionals and agreed on the usefulness of delay as an audit filter in trauma care. 

A similar <!-- Similar to what?-->German study [@horster2020empirical] looked at the correlation
between audit filters and mortality on top of a interdiciplinary expert
review according to the QUALIFY<!-- What is this?--> process. They looked at two methods in
assessing the performance of audit filters: The indicator approach and
the clinical approach. The indicator approach in assessing performance
of audit filters was to assess if the desired indicator level was
associated with a lower mortality compared to without the desired level.
The clinical approach compared different hospitals and ranked them as
"good hospitals or "bad hospitals". Good hospitals had a lower 5-year
mortality rate among a selected patient group with quality indicator
parameters of interest compared to the RISK II predicted mortality.
Evaluation of each approach ranked the usefulness of each audit filter
between "strong association with mortality", "unclear association with
mortality" and "no association with mortality". The expert review was
carried out in the same fashion as Bieler et. al. Only 3 audit filters
was similar to the audit filters used in Karolinska university hospital:
"prehospital intubation in patients with GCS \< 9", "Time to whole body
CT", "time to first emergency procedure". Prehospital intubation showed
an unclear correlation to mortality in the clinical approach, a strong
association in the indicator approach and scored high by the review
panel. In this study "GCS \< 9 and not intubated" showed no association
with OFI. "Time to whole body CT" showed unclear association with
mortality in both approaches but scored high in expert review. "more
than 30 min until first CT" was in our case associated with OFI<!-- Careful here, was the performance really that good?-->. "time
to first emergency procedure" had no correlation with mortality but
expert review agreed that it was still useful. "\>60 min until first
intervention" was in our case associated with OFI. It is important to
note that OFI in our case was not only a reflection of mortality but a
much broader concept encompassing different aspects of care quality. It
is therefore not a comparison that holds much weight<!-- Well, you've just spent the longest paragraph in your thesis elaborating on this comparison, so it would be a shame if it didn't hold much weight... I'm just pulling your leg, but see if there's a way you can rephrase this, or consider removing it-->. If anything it
highlights the difference between hospitals in what audit filters are in
use and how they are being assessed.

The audit filters "Dead at 30 days" and "CPR and thoracotomy" showed no difference from
randomness in predicting OFI, potentially due to severe traumatic injuries inevitably
result in death within 30 days of admission to the hospital resulting in a smaller time frame for mistakes to be made.


Albaaj et. al [@albaaj2023patient] looked at patient factor associated with OFI,  identifying delays of 30-60 minutes until the first CT, higher levels of care and higher ISS as linked to OFI. Similarly they
found no clear association between 30 day survival and OFI. This could partly explain the association
between "ISS \> 15 and no team activation" and OFI prediction and
similarly regarding "ISS \> 15 and not in intensive care unit".

Sanddal et. al [@sanddal2011analysis] highlighted that OFIs predominantly exist in the ED and relate to
airway management, fluid resuscitation, and chest injury management
The only audit filter directly associated with airway management was "GCS under 9 and not intubated" which performed poorly in our case. This could suggest that an alternative audit filter related to airway management should be considered in the future.

The sole purpose of audit filters is to elevate the quality of care. Audit filters should therefore reflect the current quality gap. However when quality have been improved, these audit filters loose their purpose which can result in their performance dropping. This could suggest that rather than the audit filters simply being bad, they have actually fulfilled their purpose. It is therefore crucial to continuously update the filters. 

## Generalizability 

Given the variability in screening for OFI, what an OFI consists of and what specific audit filters are used at a given trauma center the results will vary too. Furthermore, audit filters are used to fill quality gaps and due to individual trauma centers suffering from unique problems, there will not be a golden standard set of audit filters that work everywhere.

Audit filters and quality improvement programs are however widely used and the findings in this study suggest that some audit filters perform better than others. The audit filters performing better than coincidence might be a beneficial addon to already existing quality improvement program or work as a guidance in fine-tuning audit filters already in use to suit the hospitals quality gap. However, due to the study being conducted on a single, trauma level I centre in Stockholm it is unfair to assume applicability of these findings at other hospitals in other countries. It is already suggested that the efficacy of audit filters vary depending on country. Low and middle income countries seem to benefit from certain audit filters that high income countries do not benefit from and vice versa [@berg2022perceived].  

## Strengths

As stated in the method section, some audit filters whose criteria were possible to be retrieved from the registry was manually created instead of relying on the selection process by the specialized nurse. This was done in order to minimize misclassification bias.

Further more there has been very limited high quality evidence supporting audit filter efficacy [@evans2009audit]. This study adds to the evidence suggesting that some audit filters perform better than others. 

Previous studies focus primarily on audit filters effects on reducing mortality. In this study the morbidity of patient cases is also taken into account using OFI, as an umbrella term consisting of everything from bad documentation to preventable death. 

From 2017 all trauma patients was included in the registry. 


## Limitations

Opportunity for improvement, while dichotomously defined,
includes a diverse set of outcomes. This makes it difficult to create audit filters that do well in predicting OFI.

The current screening system for OFI might also introduce bias since the
filters would favour the identification of some, but not all, errors.

The way the audit filters performance were individually calculated, as
stated in the method section, was to look at every patient case with or
without an OFI and simultaneously flagged or not flagged with the audit
filter. Consequently, even if an audit filter demonstrated a true positive correlation with an OFI, instances where the same patient case had multiple audit filters flagged introduce a degree of uncertainty regarding which specific audit filter accurately predicted the outcome.

The number of missing values varied greatly among the audit filters. As stated above, missing values from each audit filter was coded as not having that audit filter. This means that in some instances where patient cases not flagged, might actually meet the audit filter criteria, but for some reason not documented. This reduces the validity of some audit filters who displayed high amounts of missing data.

## Future Studies 

Given the expanding field of
artificial intelligence and machine learning, newer studies should focus
on using these tools to screen for patient cases at risk of
an OFI. This will reduce misclassification and ensure constant and continued good performance due to its ability to learn from the current context of trauma care. This approach is new but not unexplored. A preprint is currently available by Attergrim et. al [@attergrim2023predicting] exploring the idea of machine learning models as screening tool for potential OFIs. They found that machine learning models outperformed the currently used audit filters in predicting OFI.    

further look at OFI subgroups. What kind of OFIs are overrepresented in different audit filters.<!-- I suspect these are sentences that you're planning to complete?-->

## Health Equity

It could be of interest to study the occurrence of OFI among marginalized groups of people and women in order to assess whether some groups tend to fall out from the screening method or at a higher rate falsely labelled without OFI.

# Conclusion 
<!-- 8 filters had very poor performance and 3 were useless... I recommend that you elaborate on potential reasons and alternatives in the text above.-->
Audit filters as a tool in quality improvement programs are widely used. Strong evidence supporting their efficacy in quality improvement is scarse but seem to depend on were and when they are used, in what context and for what purpose. In this study, 8 out of 11 audit filters currently in use at Karolinska University Hospital was deemed effective in predicting OFI. 3 of the audit filters did not show promising results. These results highlight the importance of continually evaluating the relevancy of certain audit filters and to fine tune audit filter criteria to better reflect the current gap in trauma care quality. 

# Acknowledgements 

I want to thank my supervisor Martin Gerdin Wärnberg
and my co-supervisors Jonatan Attergrim and Kelvin Szolnoky. I also want
to show appreciation to my good friend and colleague Anton Wasielewski
for all the support.

# Tables



# References

::: {#refs}
:::

#Appendices
