---
title: Performance of individual audit filters in predicting opportunities for improvement
  in adult trauma patients
author: "Josef Al-Khalili"
subtitle: A single-center registry-based retrospective cohort
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: 
    reference_docx: JosefAlKhaliliExJobbSeminarieVersion.docx
csl: bmcemerg.csl
editor_options:
  markdown:
    wrap: 72
bibliography: bibliography.bib
always_allow_html: yes
---

```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
source("main.R")
library(rofi)
library(dplyr)
library(ggplot2)
library(pROC)
library(gtsummary)
library(gt)
library(boot)
library(webshot2)
library(knitr)
```

# Abbreviations

OFI -- Opportunities for Improvement

CI -- Confidence Interval

GCS -- Glasgow Coma Scale

ISS -- Injury Severity Score

ED -- Emergency Department

AUC -- Area Under Curve

ROC -- Receiver Operating Characteristics

CPR -- Cardiopulmonary Resuscitation

SBP -- Systolic Blood Pressure

WHO -- World Health Organisation

YLL -- Years Life Lost

DALY -- Disability Adjusted Life Years

PTSD -- Post Traumatic Stress Disorder

# Sammanfattning

*Bakgrund:* Trauma står för 9% av globala dödsfall årligen och har en
betydande påverkan på individer och samhället.
Kvalitetsförbättringsprogram (QI) syftar till att minimera dessa
ogynnsamma utfall genom bland annat användning av granskningsfilter (AF)
och interdiciplinära Morbiditets- och Mortalitetskonferenser (MMC). AF
bygger på målskriterier för standardvård och har möjlighet att fungera
som ett screeningverktyg av relevanta patientfall för diskussion i MMC.
Ett av målen med MMC är att besluta om potentiella
förbättringsmöjligheter (OFI). *Syfte:* Denna studie syftade till att
utvärdera enskilda AFs förmåga att förutsäga OFI och att bedöma deras
relevans som screeningverktyg för MMC. *Material och Metoder:*
Registerdata från 8 309 patientfall samlades in och analyserades.
Sensitiviteten, specificiteten och Ytan under ROC-kurvanalys (AUC)
beräknades för varje enskilt AF för att utvärdera prestanda. Ett
empiriskt bootstrapping test på AUC värdena gjordes för att jämföra
prestandan med slumpmässighet (AUC = 0.5) och minimumkravet för
diagnostisk (AUC = 0.8). *Resultat:* Fyra av 11 AF hade en AUC på 0,5.
De återstående AF visade en korrelation med OFI, med AUC-värden som
varierade från 0,51 till 0,63. AF "\> 60 min till första ingrepp"
uppvisade det högsta AUC-värdet (0,63, 95% CI [0,61-0,65]). *Slutsats:*
Inget enskilt AF tycktes prestera väl i att förutsäga OFI. Detta antyder
behovet av att ompröva den nuvarande screeningmetoden för MMC.

# Abstract

*Introduction:* Trauma is responsible for 9% of global deaths annually
and has a significant impact on individuals and society. Quality
Improvement (QI) programs seek to minimize these adverse outcomes using
Audit Filters (AFs) and interdisciplinary Morbidity and Mortality
Conferences (MMC). AFs are based on criteria for standard care and can
function as a screening tool of eligible patient cases for discussion at
MMCs. One of the goals with MMCs is to determine potential Opportunities
for Improvement (OFI). *Aims:* This study aimed to evaluate the
performance of individual AFs in predicting OFIs and to assess their
relevance as a screening tool for Morbidity and Mortality Conferences
(MMC). *Material and Methods:* Register data from 8,309 patient cases
were collected for analysis. The sensitivity, specificity and Area Under
the Receiver Operating Characteristics Curve (AUC) was calculated to
assess performance. An empirical bootstrapping test of AUC values was
conducted to compare the performance of each AF to coincidence (AUC =
0.5) and minimum AUC for diagnostic tests (AUC = 0.8). *Results:* Four
out of the 11 AFs assessed had an AUC of 0.5 (p \> 0.05). The remaining
AFs demonstrated a correlation with OFI, with AUC ranging from 0.51 to
0.63. The AF "\> 60 min until first intervention" exhibited the highest
AUC (0.63, 95% CI [0.61-0.65]). *Conclusion:* No individual AF seemed to
perform well in predicting OFI, which suggest rethinking the current
screening method for MMCs.

*Keywords:* Trauma, Opportunities for Improvement, Audit Filter

# Introduction

Trauma is defined as a physical injury of external origin together with
the associated reaction of the body. This phenomenon accounts for 8-9%
of global deaths annually [@world2014injuries; @roth2018global]. Trauma
affects low- and middle-income countries at a higher rate compared to
high-income countries. In addition, young people are at higher risk of
trauma compared to the general population. It is the leading cause of
death in people between the ages of 10 and 49 [@vos2020global;
@roth2018global]. The most common type of trauma for people of all ages
is road traffic injuries [@vos2020global; @roth2018global]. In Sweden,
road traffic injuries account for approximately 40% of injuries and
falls for 40% [@SweTrau2022Arsrapport].

Trauma is broadly divided into two groups according to the underlying
causal mechanism: penetrating and blunt trauma. Commonly known examples
of penetrating trauma are gunshot wounds and knife injuries. Penetrating
trauma is often related to interpersonal violence or attempted suicide
[@stormann2016epidemiology]. Blunt trauma, on the other hand, is caused
by injuries such as road traffic accidents and falls
[@stormann2016epidemiology]. In general, both types of trauma can also
be characterised by the magnitude of the force causing the injury and
the area of impact. Blunt trauma is caused by a high force impact over a
larger area. In contrast, penetrating trauma is caused by a force
applied to a smaller area, causing greater pressure so that it is
penetrating and often results in deeper lacerations to tissues and
organs. However, blunt trauma with sufficient force can be penetrating
in nature [@hexdall2011Mechanism]. Trauma varies widely and has both
long and short term consequences on an individual, community, societal
and economic level [@newnam2014impacts].

Trauma is responsible for 11.9% of years of life lost (YLL) worldwide
[@roth2018global] and the top cause for disability-adjusted life-years
(DALY) [@vos2020global]. In addition, trauma is associated with a higher
propensity towards depression, post-traumatic stress disorder (PTSD) and
negative effects on cognitive function [@kessler1995posttraumatic;
@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts].
It is estimated that up to 11% of trauma patients have PTSD three years
after a traumatic event [@visser2017course]. In addition, people who
witness trauma are also at a higher risk of developing PTSD. Trauma is
associated with 36% not returning to work within one year
[@david2022measuring].

However, trauma is not only a burden on the person affected, but also on
communities and society [@newnam2014impacts]. One aspect of such
societal impact is the supposed high economic cost that comes with
trauma [@wesson2014cost]. Attempts have been made to estimate the global
cost of trauma, which is a complicated task due to its variability.
However, it is estimated that the cost of road traffic injuries alone
accounts for 2% of gross domestic product (GDP) in high-income countries
[@world2014injuries].

## Trauma Care Systems

Trauma care is a multidisciplinary, complex and time-critical health
care service provided by hospitals. A variety of different specialities
are involved in trauma care, including surgeons, orthopaedists,
anaesthesiologists, emergency physicians and nurses
[@American2022resources] Trauma care is divided into three components:
pre-hospital, in-hospital and post-hospital. The pre-hospital component
constitutes first responders, basic life support and advanced life
support [@world2005prehospital]. The in-hospital component is defined as
the moment a trauma patient is admitted to a health care facility, often
a trauma centre. The post-hospital component involves the rehabilitation
and recovery of treated trauma patients.

Trauma centres have a significantly lower mortality rate compared to
non-trauma centres, and the maturity of a trauma centre correlates with
lower mortality, demonstrating the importance of high-quality,
specialised trauma care [@alharbi2021effectiveness; @mann1999systematic;
@candefjord2020mortality]. Despite this, even established trauma care
systems are prone to errors due to the complex nature of trauma and its
susceptibility to delays [@nikouline2021errors; @vioque2014classifying;
@o2013opportunities], which in turn can lead to severe or even
life-threatening complications. Studies between 1990 and 2014 have shown
a pooled preventable death rate of 20%, with more recent studies showing
a lower preventable death rate [@kwon2014pooled]. For example, a recent
study in a Swedish Level I trauma centre found a preventable death rate
of 4% [@ghorbani2018analysis].

## Quality Improvement Programmes

To minimise errors in trauma care, quality improvement (QI) programs
have been established by the World Health Organization (WHO) and the
International Association for Trauma Surgery and Intensive Care (IATSIC)
[@world2009guidelines]. These programs form the core of current trauma
care systems and their development around the world
[@American2022resources].

QI programs consist of a series of recommended techniques aiming to
facilitate discussion and analysis of the trauma process in order to
take corrective action for future success. The following techniques are
commonly used: a multidisciplinary mortality and morbidity conference
(MMC), a preventable death review panel, and the use of AFs also known
as quality indicators [@world2009guidelines].

### Morbidity and Mortality Conferences

The MMC is a central QI technique. It is a proven framework with the
critical components of anonymity, focusing on specific adverse events,
recognising flawed approaches, critical analysis, improvement proposals
and implementing such [@gregor2016morbidity].

QI-centred MMCs are regular meetings broadly defined by five themes: (1)
a clear definition of the role of the MMC, (2) involvement of
stakeholders, (3) recognising and selecting appropriate patient cases
for presentation, (4) structured and focused discussions, and (5)
formulating recommendations and assigning follow-up actions
[@giesbrecht2016morbidity; @churchill2020quality].

The conference participants are healthcare professionals involved in
trauma care and its processes, but not necessarily the specific patient
case. The selection of patient cases varies, but common themes for
discussion are unexpected mortality, unexpected morbidity and errors
[@orlander2003morbidity]. It is recommended that they last about 40
minutes, but this varies between hospitals [@world2009guidelines].

Implemented effectively, MMCs serve as an important tool for error
reduction without being perceived as negative feedback for individual
care providers, but rather as an incentive for reporting. These
conferences promote a more comprehensive, systems-based approach that
emphasises task analysis, teamwork and quality improvement
[@lazzara2022morbidity]. The introduction of MMC has been shown to
increase morbidity and mortality reporting [@kong2016analysis].

### Audit Filters and Opportunities for Improvement

AFs, also known as quality indicators are established criteria based on
standardised care. Any deviation from the AFs should raise the attention
of health care professionals with the purpose of improving the quality
of care [@American2022resources]. An AF can involve outcomes such as
"death" or criteria related to guidelines and gold standards such as
"the placement of two large bore intravenous lines within 15 minutes
from arrival to a health care facility" [@world2009guidelines].

Previous studies have shown a reduction in trauma-related mortality
after the implementation of AFs [@chadbunchachai2001comparative;
@chadbunchachai2003study]. Per contra, newer studies seem to differ. In
2009, an extensive review by Evans et al. [@evans2009audit] sought to
determine if AFs could be used in improving processes in trauma care.
They found no studies successfully providing evidence for the use of
AFs. More than 1,500 AFs have been identified in a review by Stelfox et
al. [@stelfox2010quality]. Notably, they did not find a common set of
clearly defined, evidence-based, broadly accepted AFs for evaluating the
quality of trauma care. The perceived usefulness of audit filters still
remains high [@berg2022perceived]

Opportunities for improvement (OFIs) are the endpoint of patient care.
Following the review of individual patient cases there is a decision
made regarding the existence of an OFI [@American2022resources]. It has
been shown that such review processes are associated with high-quality
trauma care [@santana2014development]. OFIs are typically associated
with failures in initial care [@vioque2014classifying], specifically in
airway management, fluid resuscitation, haemorrhage control and chest
injury management [@sanddal2011analysis; @roy2017learning;
@o2013opportunities].

There is lacking evidence for the use of AFs for improving trauma care
[@evans2009audit; @stelfox2010quality] and the focus has been on
mortality as outcome measure [@chadbunchachai2001comparative;
@chadbunchachai2003study]. Furthermore, the use of AFs has also been
associated with high frequencies of false positive OFI predictions,
ranging from 24% to 80% [@sanddal2011analysis; @roy2017learning], and
some AFs do not appear to correlate with OFIs at all [@lewis2017delay].

## Aim

The aim of this study is to determine the performance of AFs in
predicting OFIs and furthermore assess their relevance as a screening
tool for MMCs.

# Methods

The research employed a single centre, registry-based, retrospective
cohort study with data from the Swedish Trauma Registry (SweTrau) and
the local trauma care quality database at Karolinska University Hospital
in Solna, to evaluate the performance of individual AFs in predicting
OFIs.

## Study setting and population

Karolinska University Hospital is classified as a Trauma level I
hospital. Every year around 1,500 trauma patients receive treatment at
the hospital. If a patient case results in team activation it is added
to the Karolinska trauma registry. If no team activation occurs but the
patient has an Injury Severity Score (ISS) of less than 9
retrospectively, it is also included in the registry. The Karolinska
trauma registry reports to SweTrau. The registry includes data on vital
signs, times, injuries, and interventions as well as patient
demographics according to the European Consensus Statement, the Utstein
template [@ringdal2008utstein]

The Karolinska trauma registry also contains a local care quality
database including specific AFs and OFIs that are determined in MMC
conferences through consensus decisions.

This process of determining OFIs unfolds in multiple stages,
characterised by escalating levels of scrutiny. Notably, instances of
mortality are directly referred to the multidisciplinary conference,
where, in addition to assessing OFIs, a decision is made regarding
whether the death was preventable or potentially preventable, a
classification also falling under the purview of OFIs.

From 2013 to 2017 there was an effort to identify adverse outcomes that
where unrelated to mortality, whereafter the review process underwent
subsequent refinement and formalisation during the study period. During
the initial period each trauma patient case underwent individual
assessments by a specialised trauma nurse to identify potential OFIs.
Only in 2017 was this procedure formalised and it became standard
practice to incorporate a preliminary individual evaluation by a
specialised trauma nurse upon data registration in the trauma registry
and the trauma quality database. The trauma quality data underwent
screening of AFs. All cases falling within the criteria delineated by
these filters, along with those trauma patients flagged by the nurse
during the initial review for possible care failures, were subjected to
a secondary review by two specialised nurses. Subsequent identification
of a potential OFI during this second review prompted a comprehensive
evaluation of the relevant trauma patient's case.

## Participants

The study included all patients treated at Karolinska University
Hospital from 2012 to 2022 who underwent screening for OFIs. Exclusion
criteria were applied to those not screened for OFI and to individuals
under the age of 15, as their clinical management process has notable
distinctions compared to those applicable to the adult population. A
total of 8,309 patients were included in the study.

## Variables

### Outcome

The outcome variable in this study is an OFI, as established by the MMCs
and furthermore treated as a dichotomous variable with "Yes - At least
one OFI identified" and "No - No OFI identified".

### Exposures

The 11 AFs used at Karolinska University Hospital function as the
exposure variable. These filters can be seen in Table 1. Each audit
filter is labelled as either "original" or "manually created" depending
on how they were managed. Manually created AFs were calculated using
trauma registry data instead of relying on the decision by the
specialised nurse as recorded in the quality database. The original AFs
were instead used as recorded in the quality database, since their
criteria encompassed variables outside of what could be retrieved from
the existing data in the registry.

## Data sources/measurement

The data used in this retrospective cohort was retrieved from the
Karolinska University Hospital trauma registry and trauma care quality
database during the period from 2012 to 2022.

Patient data on vital signs, care processes and interventions, level of
care and time aspects was retrieved from SweTrau while both exposures
(AFs) and outcome (OFI) were retrieved from the Karolinska University
Hospital trauma care quality database.

All data was anonymised to protect patient privacy.

## Bias

It is possible that consensus decisions from the MMC conferences are
flawed, leading to misclassification bias. Furthermore, the process of
flagging a patient case with an AF is done manually by a specialised
nurse as mentioned above. This means that there may be a selection bias.

Health care professionals who attend MMCs and have also been a witness
to a specific patient case might not remember certain key elements that
may affect the decision-making around OFI. Patient case reviews in this
fashion are subject to recall bias. On the other hand, health care
professionals who have not witnessed that particular case may be over-
or underestimating the significance of an OFI.

## Statistical methods

The study results were be generated by statistical analytics methods
with the help of the statistical programming language, R
[@R2022language].

Patient characteristics for patient cases with and without OFI were
compared using Wilcoxon signed-rank test for continuous variables and a
Chi-squared test for categorical values.

To evaluate the predictive performance of AFs, the sensitivity and
specificity of each individual AF in predicting OFI was calculated.
Additionally, a receiver operating curve (ROC) for binary classification
were constructed and the area under the curve (AUC) calculated in order
to compare the performance of each AF [@robin2011proc].

Using a bootstrapping method with 1,000 iterations of the selected data,
95% confidence intervals for sensitivity, specificity and AUC were
calculated. The significance of the AUC of each AF was then compared
with that of coincidence (AUC = 0.5) and the minimum standard for
diagnostic tests (AUC = 0.8) [@nahm2022receiver] through an empirical
bootstrap test with 1,000 iterations.

Instances of missing values for AFs were treated as negative, as in "not
flagged".

## Ethical considerations

In Sweden register-based studies do not require informed consent by
patients in the registry. Approval by the Swedish Ethical Review
Authority is required in order to execute a study in this manner. This
authority comprises of medical, judicial and lay members appointed by
the government. In order to uphold the principle of autonomy, measures
are in place to safeguard patient dignity.

All patient data is anonymised according to General Data Protection
Regulation (GDPR) standards [@gdpr2016EU]. The registry used provide
data on various factors such as date and time of admission, patient
characteristics e.g. age and gender. Theoretically, these attributes
could make patient cases identifiable through external sources like news
reports. Nonetheless, this means that such information must be leaked,
which would only be possible through direct registry access. In order to
minimise the risk of a potential data breach, the registry is password
protected and access to the registry server is only possible while
connected to a VPN with special credentials requiring two-factor
authentication.

This was a retrospective cohort study and as such, no patients were put
through any additional interventions and all patients were given care
according to standard. Thus, patients were at no risk of any harm due
the study.

This study will possibly give new insights to current quality
improvement programs by assessing the advantage of AFs. These new
insights, if implemented, might further lead to a better outcome for
trauma patients in terms of reduced mortality and morbidity.

This study was approved by Stockholm Research Ethics Review Board,
approval number 2021-02541 and 2021-03531.

Table 1: All 11 AFs used at Karolinska University hospital and included
in this study. It shows which AFs were created manually by registry data
(manually created) and which AFs were used as recorded in the trauma
care quality database (original).

# Results

Figure 1 presents a flowchart showcasing the exclusion and inclusion of
the patient cases from the trauma registry. After excluding patients
under 15 years old and patients not screened for OFIs, a total of 8,309
individuals were included in the study. 7,797 out of 8,309 (93.8%)
patient cases did not have an OFI and 512 out of 8,309 (6.2%) had an
OFI.

## Missing Data

Table 2 shows the missing data for each AF. As mentioned in the method
section, these values were interpreted as "not flagged" with an AF. The
variables with the highest amount of missing data were "time to
definitive treatment" (n = 5,990, 72.1%) and "time to first CT" (n =
1,012, 12.2%). The smallest amount of missing data was found in
"SBP\<90" (n = 13, 0.2%).

## Participant Characteristics

Table 3 presents the baseline characteristics of the study participants,
comparing those with and without OFI. The variables presented in Table
3, except gender and age, were used in the "calculation"manually
created" AFs showed in Table 2.

Most trauma patients were men (n = 5,755, 69%). Patient characteristics
varied depending on whether patients had an OFI or not. Respectively,
the mean age was significantly higher\
`r inline_text(tableOne, variable = pt_age_yrs, column = stat_2)` vs
`r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)`,
`r inline_text(tableOne, variable = pt_age_yrs, column = "p.value")` and
the ISS was significantly higher
`r inline_text(tableOne, variable = ISS, column = stat_2)` vs
`r inline_text(tableOne, variable = ISS, column = stat_1)`,
`r inline_text(tableOne, variable = ISS, column = "p.value")`. The
number of intubations in the emergency department was significantly
higher
`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_2)`
vs
`r inline_text(tableOne, variable = ed_intubated, level = "No", column = stat_1)`,
`r inline_text(tableOne, variable = ed_intubated, level = "No", column = "p.value")`.
Patients with OFI had longer times to definitive treatment from hospital
arrival compared to patients without OFI
`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2)`
vs
`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_1)`,
`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = "p.value")`.
The time to the first CT was also longer in patients with OFIs compared
to patients without
`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_2)` vs
`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_1)`,
`r inline_text(tableOne, variable = dt_ed_first_ct, column = "p.value")`.

The number of OFIs were highest in patients admitted to the intensive
care unit
`r inline_text(tableOne, variable = host_care_level, level = "Intensive care unit", column = stat_2)`
and surgical ward
`r inline_text(tableOne, variable = host_care_level, level = "Surgical ward", column = stat_2)`.

Gender `r inline_text(tableOne, variable = Gender, column = "p.value")`,
the mean systolic blood pressure
`r inline_text(tableOne, variable = ed_sbp_value, column = "p.value")`,
death at 30 days
`r inline_text(tableOne, variable = res_survival, column = "p.value")`
and trauma team activation
`r inline_text(tableOne, variable = Tr_Nivå, column = "p.value")` was not
significantly correlated with OFI.

Notably, there was a statistical significant correlation between GCS at
ED and OFI (p \< 0.001).

## Performance of Individual Audit Filters

The performance of each AF was determined through a calculation of
sensitivity and specificity which is demonstrated in Table 4.

The number of patient cases flagged by each individual AF varied between
107 for "GCS\<9 and not intubated" and 3907 for "\>30 min until CT". The
AF with the highest sensitivity was "\>30 min until CT" (59.2%, 95% CI
[55.2-63.8]). The lowest sensitivity was seen in "GCS\<9 and not
intubated" (1.2%, 95% CI [0.2-2]). The AF with the highest specificity
was "GCS\<9 and not intubated" (98.7%, 95% CI [98.5-98.9]). The lowest
specificity was seen in "\>30 min until CT" (53.8%, 95% CI [52.7-54.8]).
Notably, the confidence intervals varied between sensitivity and
specificity, with sensitivity displaying a greater interval.

Table 5 summarises the performance of each individual AF in identifying
OFIs in trauma care by AUC. Four of the 11 AFs, "SBP\<90", "dead at 30
days", "GCS\<9 and not intubated" and "CPR and thoracotomy" showed a
performance in predicting OFIs equal to that of coincidence (AUC = 0.5).
The highest AUC was seen in "\>60 min until first intervention" (0.63,
95% CI [0.61-0.65]). All AFs showed significantly lower AUC compared to
the minimum standard for diagnostic tests (AUC = 0.8) (p \< 0.001).

Table 2: Comparison of missing values for each AF. Missing values is
shown as a number and percentage of the total number of patient cases
included in the study (n = 8309).\
`r tableTwo` Table 3: Patient characteristics of individuals with and
without an OFI. Missing values are shown only when they exist.
`r tableOne` Table 4: Comparison of performance for each AF in
predicting OFIs. Performance is defined by sensitivity (%) and
specificity (%). A 95% confidence interval is shown together with each
performance indicator. The number of cases flagged by each filter is
displayed.\
`r tableThree` Table 5: Comparison of performance for each AF in
predicting OFIs. Performance is defined by the area under the receiver
operating characteristics curve. P-values show the significance of AUC
compared to coincidence (AUC = 0.5) and to the minimum standard for
diagnostic tests (AUC = 0.8) [@nahm2022receiver]. The p-values were
calculated using an empirical bootstrap test. Furthermore, a 95%
confidence interval was used. `r tableFour`

```{r fig1, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE, error=FALSE}
knitr::include_graphics("ofi-flowchart.pdf")
```

Figure 1: Flowchart of the patient selection process. Shows the patients
excluded and included in this study.

# Discussion

The aim of this study was to evaluate the performance of 11 individual
AFs in predicting OFIs and to assess their suitability as a screening
tool for MMCs. Most importantly, this study showed that AFs do not
perform well in predicting OFIs.

## Interpretation of Results

The observed twofold prevalence of male trauma patients compared to
females is consistent with previous findings both nationally
[@SweTrau2022Arsrapport] and globally [@roth2018global]. This is
believed to lie in the tendency toward high-risk behaviours in males.

A previous study conducted at Karolinska University Hospital by Albaaj
et al. [@albaaj2023patient] sought to identify patient factors
associated with OFIs and found that a higher ISS, longer times to CT,
longer times to first intervention and higher levels of care was
correlated with OFIs. These findings correspond with our analysis of
patient characteristics and AFs linked to these factors.

Higher ISS and higher levels of care in patients with OFIs as opposed to
patients without OFIs could be explained by the greater complexity of
such trauma cases. Greater complexity requires more interventions which
could allow for more mistakes [@albaaj2023patient].

Notably, GCS at ED showed a correlation with OFI even though the median
GCS and its IQR was identical for individuals with OFI and without. Upon
further analysis this could be explained by the difference in
distribution outside the IQR.

AFs generally exhibited a high specificity relative to their sensitivity
which may be attributed to the outcome being rather uncommon, whilst
every individual AF misses the majority of OFIs. This could suggest that
the AFs are too rigid in their criteria. On the contrary, a too
inclusive criteria might flag several patients, who do not necessarily
have an OFI [@willis2008interpreting].

In 2008, Willis et al. [@willis2008interpreting] assessed 14 audit
filters, finding only three to be associated with an increased risk of
poor outcomes. One AF, "GCS\<9 and not intubated", also used in this
study, similarly showed no significant ability to predict OFIs. An
explanation for this could be that an abnormal GCS as a result from
intoxication is not comparable to an abnormal GCS due to head injury for
example, suggesting that the AF might be more reflective of clinical
decision rather than quality of care [@willis2008interpreting]. However,
according to Sanddal et al. [@sanddal2011analysis] airway management is
associated with OFI suggesting that an alternative audit filter related
to airway management might potentially perform differently.

Both AFs measuring delay: "\>60 until first intervention" and "\>30min
until first CT", showed a correlation with OFI. This is comparable to
what Teixeira et al. [@teixeira2007preventable] demonstrated, with delay
correlating to preventable death. Furthermore, Ghorbani et al.
[@ghorbani2018analysis] identified delay in treatment, delay to CT, and
clinical judgement errors as common errors in trauma care. A more recent
study in Japan by Yamamoto et. al [@yamamoto2023immediate] showed a
significant reduction in mortality when patients received a whole-body
CT within 10 minutes of arrival at the ED. It is important to note that
OFI in this study was not only a reflection of mortality but a broader
concept encompassing different aspects of care quality. If anything, it
highlights the difference between hospitals, which AFs are in use, and
how they are being assessed.

AFs that present with a high risk of inevitable death showed poor"Dead
at 30 days" and "CPR and thoracotomy", could be explained by inevitable
mortality, thus presenting with a smaller time frame for mistakes to be
made.

## The audit filter dilemma

AFs are static in their criteria, contrary to health care in general.
Consequently, the utility of AFs may diminish if confronted with
emerging challenges or upon resolution of the quality gaps they were
designed for. The static nature of AFs highlight their need for
consistent and evidence-based review to determine current usefulness
[@santana2014development]. However, it is worth pointing out that the
review process might present with costs greater than the potential
benefit from AFs.

AFs face noteworthy challenges due to over-reliance, bias and
confounders. Firstly, while AFs may help in determining the existence of
OFI, they could overshadow other factors that may contribute to that
analysis, leading to over-reliance. Secondly, devising AFs comes with an
unintended bias. They might disproportionally affect certain groups of
people failing to recognise patient cases that fall outside the scope of
AF criteria [@willis2008interpreting]. For example, AFs primarily focus
on aspects of trauma care mostly relevant to medical doctors despite the
interdisciplinary nature of trauma care. This is especially
counterintuitive given the broad definition of OFIs. Lastly, confounding
variables adds to the uncertainty of AFs performance. For example,
patients receiving the same interventions at different time points may
do so as a function of injury severity not quality of care
[@willis2008interpreting].

## Generalizability

AFs are widely used but generalizability is complicated due to the lack
of common grounds on what they are, what purpose they serve and how they
shall be used [@stelfox2010quality; @stelfox2011evidence;
@berg2022perceived].

Our findings suggest that AFs perform poorly in predicting OFIs. The AFs
with performance exceeding coincidence (AUC = 0.5) might be a beneficial
addition to an already existing quality improvement programme or could
be used to fine-tune AFs already in use to plug hospitals' quality gaps.
Due to the study being conducted in a single, trauma level I centre in
Stockholm it is unrealistic to assume applicability of these findings at
other hospitals, in other countries, facing unique challenges
[@world2009guidelines]. For example, Low and middle income countries
seem to benefit from certain AFs that high income countries do not and
vice versa [@berg2022perceived].

## Strengths

As stated in the method section, some AFs were manually created from
registry data instead of relying on the selection process by the
specialised nurse (see table 1). This, in order to minimise
misclassification bias.

Furthermore, there has been very limited high quality evidence
supporting the efficacy of AFs in improving trauma care
[@evans2009audit; @stelfox2010quality] and predicting OFI
[@sanddal2011analysis; @roy2017learning]. This study shows comparable
results. Previous studies also focus primarily on AFs' ability to reduce
mortality [@chadbunchachai2001comparative; @chadbunchachai2003study]. In
this study the morbidity of patient cases is also taken into account
using OFIs as an umbrella term that includes factors ranging from bad
documentation to preventable death.

From 2017 all trauma patients admitted to Karolinska University Hospital
were included in the registry reducing the risk of bias in inclusion of
patient cases.

## Limitations

Opportunity for improvement, while dichotomously defined, includes a
diverse set of outcomes. This convolutes the process of creating AFs
that perform well in predicting OFIs. The current screening system for
OFIs might also introduce bias since the AFs would favour the
identification of some, but not all, errors.

Each AF was individually related to OFI. Consequently, even if an AF
demonstrated a true positive correlation with an OFI, instances where
the same patient case had multiple AFs introduce a degree of uncertainty
regarding which specific AF accurately predicted the outcome.

The number of missing values varied greatly among the AFs. As stated
above, missing values from each AF were coded as "not flagged" by an AF.
This means that some instances where patient cases were not flagged,
might actually meet the AF criteria, but for some reason not documented.
This reduces the validity of some AFs that featured high amounts of
missing data.

## Future Studies

With the ongoing academic, political, and public attention towards
patient safety and quality of healthcare [@evans2009audit], it is
important to discover optimal methods for monitoring and enhancing the
care provided within trauma systems.

Given the expanding field of artificial intelligence and machine
learning, newer studies should focus on using these tools to screen for
patient cases at risk of OFIs. This will diminish misclassification
rates and ensure consistent and sustained high performance, attributable
to an algorithm's capability to adapt and learn from the prevailing
context of trauma care. This approach is contemporary but not
unexplored. A preprint is currently available by Attergrim et al.
[@attergrim2023predicting] exploring the idea of machine learning models
as a screening tool for potential OFIs. They found that machine learning
models outperformed the AFs currently used at Karolinska University
Hospital in predicting OFIs.

Should the use of AFs remain, it is important to ensure its performance
by routinely updating them in order to better reflect the current
quality gaps.

## Health Equity

There is a growing body of research [@schoenfeld2013patient;
@bedri2017national] suggesting that certain populations have poorer
health and treatment outcomes due to implicit biases that may affect
screening and treatment by health care professionals, sometimes leading
to unequal distributions of mortality and morbidity among minority
groups. Reviewing OFIs in adult trauma patients from this perspective
could help uncover and address health inequities in this regard. Since
there is a gender disparity between the number of trauma patients, it
could be worthwhile to look further into whether gender disparities in
OFIs could reveal gender discrimination in the screening and labelling
of OFIs.

# Conclusion

AFs are widely used as a tool in quality improvement programmes and
should reflect the current gap in trauma care quality. Strong evidence
supporting their efficacy in quality improvement is scarce. In this
study, no audit filter seemed to show promising performance in
predicting OFIs with AUC ranging from 0.5 to 0.63. These results
highlight the importance of continually evaluating the relevance of AFs.
However, this process takes time and resources and using an alternative
to AFs is recommended.

# Acknowledgements

I wish to thank my supervisor Martin Gerdin Wärnberg and my
co-supervisors Jonatan Attergrim and Kelvin Szolnoky for their support
and guidance throughout the entirety of this study.

#Relevant?

An extensive review by Evans et al. [@evans2009audit] sought to review
the evidence for audit filters as a tool to improve care. A total of 741
studies were screened by title. Of these studies, 42 abstracts were
appraised and five complete articles were examined in detail. None of
the five studies met the inclusion criteria established. Two studies by
Chadbunchachai et al. [@chadbunchachai2001comparative;
@chadbunchachai2003study], which were examined in detail by Evans et. al
were mentioned in WHO:s guidelines for QI programmes
[@world2009guidelines].They were excluded from the Evans et al. review
due to no clear pre‐ or post‐intervention data points or a clearly
defined time when the respective interventions were implemented. This
suggests that audit filters might not serve their purpose.

Audit filters demonstrating a performance equal to coincidence are
deemed unsuitable as screening tools for OFIs at Karolinska University
Hospital. On the contrary, audit filters demonstrating a performance
superior to that of coincidence could still be used as audit filters.
However, it is important to note that despite these audit filters
showing correlation with OFI, the AUC was lower than the minimum for
diagnostic tests (AUC = 0.8) [@nahm2022receiver].

# References

::: {#refs}
:::
