---
title: Performance of individual audit filters in predicting opportunities for improvement
  in adult trauma patients
author: "Josef Al-Khalili"
subtitle: A single-center registry-based retrospective cohort
output:
  word_document: 
    reference_docx: JosefAlKhaliliExJobbSeminarieVersion.docx
  pdf_document: default
  html_document:
    df_print: paged
csl: bmcemerg.csl
editor_options:
  markdown:
    wrap: 72
bibliography: bibliography.bib
always_allow_html: yes
---

```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
source("main.R")
library(rofi)
library(dplyr)
library(ggplot2)
library(pROC)
library(gtsummary)
library(gt)
library(boot)
library(webshot2)
library(knitr)
```

# Abbreviations

OFI -- Opportunities for Improvement

CI -- Confidence Interval

GCS -- Glasgow Coma Scale

ISS -- Injury Severity Score

ED -- Emergency Department

AUC -- Area Under Curve

ROC -- Receiver Operating Characteristics

CPR -- Cardiopulmonary Resuscitation

SBP -- Systolic Blood Pressure

WHO -- World Health Organisation

YLL -- Years Life Lost

DALY -- Disability Adjusted Life Years

PTSD -- Post Traumatic Stress Disorder

# Sammanfattning

*Bakgrund:* Trauma står för 9% av globala dödsfall årligen och har en
betydande påverkan på individer och samhället.
Kvalitetsförbättringsprogram (QI) syftar till att minimera dessa
ogynnsamma utfall genom användning av audit filter som indikatorer för
standardvård och genom att utvärdera patientfall för potentiella
förbättringsmöjligheter (OFI). *Syfte:* Denna studie syftade till att
utvärdera prestandan hos audit filter för att förutsäga OFI och att
bedöma deras relevans som screeningverktyg för Morbiditets- och
Mortalitetskonferenser (MMC). *Material och Metoder:* Sensitivitet,
specificitet och Area Under the Receiver Operating Characteristics Curve
(AUC) beräknades för att utvärdera prestanda. En t-test för AUC
genomfördes för att jämföra prestanda med slumpmässighet. *Resultat:* Av
de 11 audit filter som utvärderades hade 3 en AUC på 0,5 (p \> 0,05). De
återstående audit filterna visade en korrelation med OFI, med AUC-värden
som varierade från 0,51 till 0,63. Audit filtret "\> 60 min till första
ingrepp" uppvisade den högsta AUC på 0,63. *Slutsats:* Ingen enskild
revisionsfilter tycktes prestera väl i att förutsäga OFI. Detta antyder
behovet av att ompröva den nuvarande screeningmetoden för MMC.

# Abstract

*Introduction:* Trauma is responsible for 9% of global deaths anually
and has a significant impact on individuals and society. Quality
Improvement (QI) programs seek to minimize these adverse outcomes
through the use of audit filters as indicators of standard care and
through assessing patient cases for potential Opportunities for
Improvement (OFI). *Aims:* This study aimed to evaluate the performance
of audit filters in predicting OFI and to assess their relevance as a
screening tool for Morbidity and Mortality Conferences (MMC). *Material
and Methods:* The sensitivity, specificity and Area Under the Reciever
Operating Characteristics Curve (AUC) was calculate din order to assess
performance. A t-test for AUC was conducted to compare the performance
to coincidence. *Results:* Of the 11 audit filters assessed, 3 had an
AUC of 0.5 (p \> 0.05). The remaining audit filters demonstrated a
correlation with OFI, with AUC ranging from 0.51 to 0.63. The audit
filter "\> 60 min until first intervention" exhibited the highest AUC of
0.63. *Conclusion:* No individual audit filter seemed to perform well in
predicting OFI, which suggest rethinking the current screening method
for MMC.

*Keywords:* Trauma, Opportunities for Improvement, audit filter,
performance

# Introduction

Trauma is defined as a physical injury of external origin together with
the associated reaction of the body. This phenomenon accounts for 9% of
global deaths annually [@world2014injuries]. Trauma affects low- and
middle-income countries at a higher rate compared to high-income countries. In addition, young people are
at higher risk of trauma compared to the general population. It is the
leading cause of death in people between the ages of 10 and 49
[@vos2020global; @roth2018global]. The most common type of trauma for
people of all ages is road traffic injuries [@vos2020global;
@roth2018global]. In Sweden, road traffic injuries account for
approximately 40% of injuries and falls for 40%
[@SweTrau2022Arsrapport].

Trauma is broadly divided into two groups according to the underlying
causal mechanism: penetrating and blunt trauma. Commonly known examples
of penetrating trauma are gunshot wounds and knife injuries. Blunt trauma,
on the other hand, is caused by injuries such as road traffic accidents
and falls. In general, both types of trauma can also be characterised
by the magnitude of the force causing the injury and the area of impact.
Blunt trauma is caused by a high force impact over a larger area. In
contrast, penetrating trauma is caused by a force applied to a smaller
area, causing greater pressure so that it is penetrating and often
results in deeper lacerations to tissues and organs. However, blunt
trauma with sufficient force can be penetrating in nature
[@hexdall2011Mechanism]. Trauma varies widely and has both long and
short term consequences on an individual, societal and economic level.

Trauma is responsible for 11.9% of years of life lost (YLL) worldwide
[@roth2018global]. In addition, trauma is associated with a higher
propensity towards depression, post-traumatic stress disorder (PTSD) and
negative effects on cognitive function [@north1999psychological;
@muscatelli2017prevalence].

It is estimated that up to 11% of trauma patients have PTSD three years
after a traumatic event [@visser2017course]. In addition, people who
witness trauma are also at a higher risk of developing PTSD. Trauma is
associated with 36% not returning to work within one year
[@david2022measuring].

However, trauma is not only a burden on the person affected, but also on
society as a whole [@newnam2014impacts]. One aspect of such societal
impact is the economic cost. Attempts have been made to estimate the
global cost of trauma, which is a complicated task due to its
variability. However, it is estimated that the cost of road traffic
injuries alone accounts for 2% of gross domestic product (GDP) in
high-income countries [@world2014injuries].

## Trauma Care Systems

Trauma care is a multidisciplinary, complex and time-critical health
care service provided by hospitals. A variety of different specialities
are involved in trauma care, including surgeons, orthopaedists,
anaesthesiologists, emergency physicians and nurses. Trauma care is
divided into three components: pre-hospital, in-hospital and
post-hospital. The pre-hospital component constitutes first responders,
basic life support and advanced life support.[@world2005prehospital].
The in-hospital component is defined as the moment a trauma patient is
admitted to a health care facility, often a trauma centre. The
post-hospital component involves the rehabilitation and recovery of
treated trauma patients.

Trauma centres have a significantly lower mortality rate compared to
non-trauma centres, and the maturity of a trauma centre correlates with
lower mortality, demonstrating the importance of high-quality,
specialised trauma care [@alharbi2021effectiveness;
@mann1999systematic]. Similarly in Sweden
[@candefjord2020mortality] showed that the 30-day adjusted risk of death is
41% lower in a mature trauma centre compared to a non-trauma centre.
Despite this, even established trauma care systems are prone to error
due to its complexity and susceptibility to delays
[@nikouline2021errors; @vioque2014classifying], which in turn can lead
to complications and death. Studies between 1990 and 2014 have shown a
pooled preventable death rate of 20%, with more recent studies showing
less [@kwon2014pooled]. For example, a recent study in a Swedish Level I
trauma centre found a preventable death rate of 4%
[@ghorbani2018analysis].

## Quality Improvement

To minimise errors in trauma care, quality improvement (QI) programs
have been established by the World Health Organization (WHO) and the
International Association for Trauma Surgery and Intensive Care (IATSIC)
[@world2009guidelines]. These programs form the core of current trauma
care systems and their development around the world.

QI programs consist of a series of recommended techniques aimed at
facilitating discussion and analysis of the trauma process in order to
take corrective action for future success. The following techniques are
used: A multidisciplinary mortality and morbidity conference (MMC), a
preventable death review panel, and the use of audit
filters[@world2009guidelines].

### Morbidity and Mortality Conferences and Preventable Death Review Panels

<!-- ändra ordning -->

The MMC is a central QI technique. It is a proven framework with the
critical components of anonymity, focusing on specific adverse events,
recognising flawed approaches, critical analysis, improvement proposals and
implementing such [@gregor2016morbidity].

MMCs are regular meetings at which specific patient cases are presented
for discussion. The conference participants are healthcare professionals
involved in trauma care and its processes, but not necessarily the
specific patient case. The selection of patient cases varies, but common
themes for discussion are unexpected mortality, unexpected morbidity and
errors [@orlander2003morbidity]. It is recommended that they last about
40 minutes, but this varies from hospital to hospital
[@world2009guidelines].

Five themes have been proposed that broadly define QI-centred MMC: (1) a
clear definition of the role of the MMC, (2) involvement of
stakeholders, (3) recognising and selecting appropriate patient cases
for presentation, (4) structured and focused discussions, and (5)
formulating recommendations and assigning follow-up
actions [@giesbrecht2016morbidity; @churchill2020quality].

Implemented effectively, MMCs serve as an important tool for error
reduction without being perceived as negative feedback for individual care
providers, but rather as an incentive for reporting. These conferences
promote a more comprehensive, systems-based approach that emphasises
task analysis, teamwork and quality improvement [@lazzara2022morbidity].
The introduction of MMC has been shown to increase morbidity and
mortality reporting [@kong2016analysis].

### Audit Filters and Opportunities for Improvement

Audit filters, also known as quality indicators are specific established
criteria used in trauma processes and care. They are used to detect
deviations from standardised care in order to further analyse the cause
of complications in trauma patients. An audit filter can involve outcomes such as "death" or criteria related to guidelines and gold standards such as
"the placement of two large bore intravenous lines within 15 minutes from
arrival to a health care facility" [@world2009guidelines]. The purpose of
audit filters is to improve the quality of care.

Previous studies have shown a reduction in trauma-related mortality after the
implementation of audit filters [@chadbunchachai2001comparative]. Per
contra, newer studies seem to differ. In 2009, an extensive
review that sought to determine if audit filters could be used in
improving processes in trauma care. None of the studies met
established inclusion criteria [@evans2009audit]. The use of audit
filters has also been associated with high frequencies of false
positives, ranging from 24% to 80%.[@sanddal2011analysis];
[@roy2017learning].

Opportunities for improvement (OFIs) are the endpoint of patient care.
Following the review of individual patient cases there is a consensus
decision made regarding the existence of an OFI. It has been shown that
such review processes are associated with high-quality trauma care
[@santana2014development]. OFIs are typically associated with failures
in initial care [@vioque2014classifying], specifically in airway
management, fluid resuscitation, haemorrhage control and chest injury
management [@sanddal2011analysis]; [@roy2017learning;
@o2013opportunities].

Audit filters can provide guidance in determining OFI which in turn can serve
as basis for developing new audit filters. There seems to be a lack of
evidence for the use of audit filters in predicting OFI
[@evans2009audit] and some audit filters do not appear to correlate with OFIs
at all [@lewis2017delay].

## Aim

The aim of this study is to determine the performance of audit filters in
predicting OFIs and furthermore assess their relevance as a screening
tool for MMC.

# Methods

The research employed a single centre, registry-based, retrospective cohort study
with data from the Swedish Trauma Registry (SweTrau) and the local
trauma care quality database at Karolinska University Hospital in Solna, to evaluate the performance of individual audit filters in
predicting OFIs.

## Study setting and population

Karolinska University Hospital is classified as a Trauma level I
hospital. Every year around 1,500 trauma patients receive treatment at
the hospital. If a patient case results in team activation it is added
to the Karolinska trauma registry. If no team activation occurs but
the patient has an Injury Severity Score (ISS) of less than 9
retrospectively, it is also included in the registry. The Karolinska
trauma registry reports to SweTrau. The registry includes data on
vital signs, times, injuries, and interventions as well as patient
demographics according to the European Consensus Statement, the Utstein
template [@ringdal2008utstein]

The Karolinska trauma registry also contains a local care quality
database including specific audit filters and OFIs that are determined
in MMC conferences through consensus decisions.

This process of determining OFIs unfolds in multiple stages,
characterised by escalating levels of scrutiny. Notably, instances of
mortality are directly referred to the multidisciplinary conference,
where, in addition to assessing OFIs, a decision is made regarding
whether the death was preventable or potentially preventable, a
classification also falling under the purview of OFIs.

From 2013 to 2017 there was an effort to identify adverse
outcomes that where unrelated to mortality, whereafter the review
process underwent subsequent refinement and formalisation during the
study period. During the initial period each trauma patient case
underwent individual assessments by a specialised trauma nurse
to identify potential OFIs. Only in 2017 was this procedure formalised
and it became standard practice to incorporate a preliminary individual
evaluation by a specialised trauma nurse upon data registration in the
trauma registry and the trauma quality database. The trauma quality data
underwent screening of audit filters. All cases falling within the
criteria delineated by these filters, along with those trauma patients
flagged by the nurse during the initial review for possible care
failures, were subjected to a secondary review by two specialised
nurses. Subsequent identification of a potential OFI during this second
review prompted a comprehensive evaluation of the relevant trauma
patient's case.

## Participants

The study included all patient cases who underwent screening for
OFIs. Exclusion criteria were applied to those not screened for OFI and
to individuals under the age of 15, as their clinical management process
has notable distinctions compared to those applicable to the adult
population.

## Variables

### Outcome

The outcome variable in this study is an OFI, as established by the MMC
conference and furthermore treated as a dichotomous variable with "Yes -
At least one OFI identified" and "No - No OFI identified".

### Exposures

The 11 audit filters used at Karolinska University Hospital function as
the exposure variable. These filters can be seen in Table 1. Each audit filter is labelled as either
"original" or "manually created" depending on how they were managed.
Manually created audit filters were calculated using trauma registry
data instead of relying on the decision by the specialised nurse as
recorded in the quality database. The original audit filters were
instead used as recorded in the quality database, since their criteria
encompassed variables outside of what could be retrieved from the
existing data in the registry.

## Data sources/measurement

The data used in this retrospective cohort was retrieved from the
Karolinska University Hospital trauma registry and trauma care quality
database during the period from 2012 to 2022.

Patient data on vital signs, care processes and interventions, level of
care and time aspects was retrieved from SweTrau while both exposures
(audit filters) and outcome (OFI) were retrieved from the Karolinska
University Hospital trauma care quality database.

All data was anonymised to protect patient privacy.

## Bias

It is possible that consensus decisions from the MMC conferences are
flawed, leading to misclassification bias. Furthermore, the process of
flagging a patient case with an audit filter is done manually by a
specialised nurse as mentioned above. This means that there may be a
selection bias.

Health care professionals who attend MMCs and have also been a witness
to a specific patient case might not remember certain key elements that
may affect the decision-making around OFI. Patient case reviews in this
fashion are subject to recall bias. On the other hand, health care
professionals who have not witnessed that particular case may be over-
or underestimating the significance of an OFI.

## Study size

The study cohort encompasses all eligible patients treated at Karolinska
University Hospital from 2012 to 2022.
<!--number
## Statistical methods

The study results will be generated by statistical analytics methods
with the help of the statistical programming language, R
[@R2022language].

To evaluate the predictive performance of audit filters, the sensitivity
and specificity of each individual audit filter in predicting OFI will
be calculated. Additionally, a receiver operating curve (ROC) for binary
classification will be constructed and the area under the curve (AUC)
calculated in order to compare the performance of each audit filter
[@robin2011proc].

Using a bootstrapping method with 1000 iterations of the selected data,
95% confidence intervals for sensitivity, specificity and AUC will be
calculated. The significance of the AUC of each audit filter will then
be compared with that of coincidence (AUC = 0.5), through a one sample
T-test.

Instances of missing values for audit filters will be treated as
negative, as in "not flagged".

## Ethical considerations

In Sweden register-based studies do not require informed consent by
patients in the registry. Approval by the Swedish Ethical Review
Authority is required in order to execute a study in this manner. This
authority comprises of medical, judicial and lay members appointed by
the government. In order to uphold the principle of autonomy, measures
are in place to ensure integrity of each patient.

All patient data is anonymised according to General Data protection
Regulation (GDPR) standard [@gdpr2016EU]. In the registry used, there is
data on various factors such as date and time of admission, and patient
characteristics e.g. age and gender. Theoretically, these attributes
could make patient cases identifiable through external sources like news
reports. Nonetheless, this means that such information must be leaked,
which would only be possible through direct registry access. In order to
minimise the risk of a potential data breach, the registry is password
protected and access to the registry server is only possible while
connected to a VPN with special credentials requiring two-factor
authentication.

This was a retrospective cohort study and as such, no patients were put
through any additional interventions and all patients were given care
according to standard. Thus, patients were at no risk of any harm due
the study.

This study will possibly give new insights to current quality
improvement programs by assessing the advantage of audit filters. These
new insights, if implemented, might further lead to a better outcome for
trauma patients in terms of reduced mortality and morbidity.

This study was approved by Stockholm Research Ethics Review Board,
approval number 2021-02541 and 2021-03531.

# Results

Figure 1 presents a flowchart showcasing the exclusion and inclusion of
the patient cases from the trauma registry. After excluding patients
under 15 years old and patients not screened for OFIs, a total of 8309
individuals were included in the study. 7797 out of 8309 (93.8%) patient
cases did not have an OFI and 512 out of 8309 (6.2%) had an OFI.

## Missing Data

Table 2 shows the missing data for each audit filter. As mentioned in
the method section, these values where interpreted as not flagged with
an audit filter. The variables with the highest amount of missing data
was "time to definitive treatment" (n = 5990, 72.1%) and "time to first
CT" (n = 1012, 12.2%). The least amount of missing data was found in
"SBP\<90" (n = 13, 0.2%).

## Participant Characteristics

Table 3 presents the baseline characteristics of the study participants,
comparing those with and without an assigned OFI. The variables
presented in Table 3, except gender and age, were used in the
calculation of audit filters.

5755 (69%) out of all patients included in the study were men and 2554
(31%) women. Results varied depending on whether patients had an OFI or
not. The mean age was slightly higher
`r inline_text(tableOne, variable = pt_age_yrs, column = stat_2)` vs
`r inline_text(tableOne, variable = pt_age_yrs, column = stat_1)` and
the ISS was higher
`r inline_text(tableOne, variable = ISS, column = stat_2)` vs
`r inline_text(tableOne, variable = ISS, column = stat_1)`. The number
of intubations in the emergency department was higher
`r inline_text(tableOne, variable = ed_intubated, level = "Yes", column = stat_2)`
vs
`r inline_text(tableOne, variable = ed_intubated, level = "No", column = stat_1)`.
Patients with OFI had longer times to definitive treatment from hospital
arrival compared to patients without OFI
`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_2)`
vs
`r inline_text(tableOne, variable = dt_ed_emerg_proc, column = stat_1)`.
Moreover, the time to the first CT was also longer in patients with OFIs
`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_2)` vs
`r inline_text(tableOne, variable = dt_ed_first_ct, column = stat_1)`.

The mean systolic blood pressure and GCS at the emergency department did
not differ between the OFI and no OFI group.

## Individual Audit Filters

### Performance of Individual Audit Filters

The performance of each audit filter was determined through a
calculation of sensitivity and specificity. This is demonstrated in
Table 4.

Varying numbers of patient cases were flagged with each audit filter.
The audit filter flagged most frequently was "\>30 min until CT". The
least flagged audit filter was "GCS\<9 and not intubated".

The audit filter with the highest sensitivity was "\>30min until CT"
(59.2% CI: 55.3-63.3). The audit filter with the lowest sensitivity was
"GCS\<9 and not intubated" (1.2% CI: 0.1-2).

The audit filter with the highest specificity was "GCS\<9 and not
intubated" (98.7% CI: 98.5-98.9). The The audit filter with the lowest
specificity was "\>30min until CT" (53.8% CI: 52.7-55.2).

Generally, the confidence intervals varied between sensitivity and
specificity, with sensitivity having a greater interval span.

Table 5 summarizes the performance of each individual audit filter in
identifying OFIs in trauma care by AUC. Three of the ten audit filters,
"dead at 30 days", "GCS\<9 and not intubated" and "CPR and thoracotomy"
showed a performance in predicting an OFI as good as coincidence (AUC =
0.5). \>60 min until first intervention had the highest AUC.

Table 2: Comparison of missing values for each audit filter as a number
and percentage of the total amount of patient cases (n = 8309).\
`r tableTwo` Table 3: Patient characteristics of individuals with and
without an OFI. Missing values are shown only when they exist.
`r tableOne` Table 4: Comparison of performance for each audit filter in
predicting OFIs. Performance is defined by sensitivity (%) and
specificity (%). A 95% confidence interval is shown together with each
performance indicator. The number of cases flagged by each filter is
displayed.\
`r tableThree` Table 5: Comparison of performance for each audit filter
in predicting OFIs. Performance is defined by the area under the
receiver operating characteristics curve. P-values show the significance
of AUC compared to coincidence (AUC = 0.5) through a t-test. A 95%
confidence interval is used. `r tableFour`

```{r fig1, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE, error=FALSE}
knitr::include_graphics("ofi-flowchart.pdf")
```

Figure 1: Flowchart of the patient selection process. Shows the patients
excluded and included in this study.

# Discussion

## Interpretation of Results

The aim of this study was to evaluate the performance of individual
audit filters in predicting OFIs, to assess their suitability as a
screening tool for morbidity and mortality conferences.
<!-- Before you start discussing how demographics differed between groups I suggest that you summarise your key findings. What were the most important findings? That all filters performed poorly? -->

The observed twofold prevalence of male trauma patients compared to
females is consistent with previous findings both nationally
[@SweTrau2022Arsrapport] and globally [@roth2018global]. This is
believed to lie in the tendecy to high-risk behaviours in males.

The higher ISS in the patients with OFIs as opposed to patients without
OFIs could be explained by the greater complexity of trauma cases with
increasing ISS. Greater complexity allows for more mistakes. These
findings stay true with a previous study conducted at Karolinska
University Hospital by Albaaj et. al [@albaaj2023patient]. Albaaj et. al
sought to identify patient factors associated with OFIs and found that a
higher ISS was significantly correlated with OFIs.

Audit filters generally exhibited a high specificity relative to their
sensitivity which may be attributed to the outcome being rather
uncommon, whilst every individual audit filter miss the majority of
OFIs. Theoretically, adding all audit filters together as one audit
filter would result in the sensitivity increasing, and simultaneously
the specificity decreasing.

An extensive review by Evans et. al [@evans2009audit] sought to review
the evidence for audit filters as a tool to improve care. A total of 741
studies were screened by title. Of these studies, 42 abstracts were
appraised and five complete articles were examined in detail. None of
the five studies met the inclusion criteria established. Two of the
studies by Chadbunchachai et. al
[@chadbunchachai2001comparative];[@chadbunchachai2003study], which were
examined in detail by Evans et. al were mentioned in WHO:s guidelines
for quality improvement programmes [@world2009guidelines]. They were
excluded from Evans et. al review due to no clear pre‐ or
post‐intervention data points or a clearly defined point in time when
the respective interventions were implemented.

In 2008, Willis et. al [@willis2008interpreting] assessed 14 audit
filters, finding only three to be associated with an increased risk of
poor outcomes. One audit filter, "GCS\<9 and not intubated", also used
in this study showed no significant ability to predict poor outcome.

Audit filters demonstrating a performance equal to coincidence are
deemed unsuitable as screening tools for OFIs. On the contrary, audit
filters demonstrating a performance superior to that of coincidence
could still be used as audit filters. However, it is important to note
that despite these audit filters proving meaningful, it does not mean
they are acceptable as a screening tool. That would not have been the
case with a diagnostics test where the AUC should be at least 0.8 for it
to be considered acceptable [@nahm2022receiver].

A noteworthy challenge with audit filters is the risk of overrelying on
them. While they may help in deciding the existence of OFI, they could
overshadow other factors that may contribute to that decision.

Audit filters are static in their criteria, oppositely to healthcare in
general. This means that audit filters may loose their purpose if new
guidelines are introduced. For example, liver and spleen injuries did
not perform at all in predicting OFIs. It could very well be that new
and improved guidelines and gold standards reduced the number of OFIs.

A further problem with audit filters being static is that they
themselves can fulfill their purpose and subsequently loose their
relevance through a negative feedback loop. For example, a problem in
health care occurs and in response an audit filter is made in order to
tackle this problem. At first this will theoretically result in a good
predictability for OFI. However, when such OFIs are taken into
consideration the audit filter will gradually loose its performance.

Some audit filter criteria are more rigid than others. A criterion too
rigid will fail to recognise the majority of OFIs while too inclusive
criteria might flag several patients, who do not necessarily have an
OFI. This in turn leads to the next problem regarding the resources and
time needed for making an OFI decision. A very inclusive criterion will
cause many patient cases being flagged and will lead to many patient
cases being discussed at MMCs.

Suggest we were to replace audit filters used by Karolinska University
Hospital with other audit filters. We would have to review them over
time and simultaneously risk the fact that they might not perform.

Another general challenge with audit filters is the unintended bias that
comes with postulating audit filters. They might disproportionally
affect certain groups of people. Moreover, the risk of confounders are
present when establishing such criteria.

Audit filters currently in use at Karolinska University Hospital focus
primarily on aspects of trauma care mostly relevant to medical doctors.
It might be beneficial to consider the interdisciplinary nature of
trauma care by using criteria related to morbidity for example.

### Delay

Both audit filters measuring delay: "\>60 until first intervention" and
"\>30min until first CT", showed a correlation with OFI. This is
comparable to what Teixeira et. al [@teixeira2007preventable]
demonstrated, with delay correlating to preventable death. Furthermore,
Ghorbani et al. [@ghorbani2018analysis] identified delay in treatment,
delay to CT, and clinical judgment errors as common errors in trauma
care, which aligns with our findings. A more recent study in Japan by
Yamamoto et. al [@yamamoto2023immediate] showed a significant reduction
in mortality when patients received a whole body CT within 10 minutes of
arrival to the ED. It is important to note that OFI in our case was not
only a reflection of mortality but a broader concept encompassing
different aspects of care quality. If anything, it highlights the
difference between hospitals, which audit filters are in use and how
they are being assessed.

The audit filters "Dead at 30 days" and "CPR and thoracotomy" showed no
difference from coincidence (AUC = 0.5) in predicting OFI, potentially
due to severe traumatic injuries inevitably result in death within 30
days of admission to the hospital resulting in a smaller time frame for
mistakes to be made.

Albaaj et. al [@albaaj2023patient] examined patient factors associated
with OFIs, identifying delays of 30-60 minutes until the first CT,
higher levels of care and higher ISS as linked to OFIs. Similarly, they
found no clear association between 30 day survival and OFIs. This could
partly explain the correlation between "ISS \> 15 and no team
activation" and OFI and similarly regarding "ISS \> 15 and not in
intensive care unit".

Sanddal et. al [@sanddal2011analysis] highlighted that OFIs
predominantly exist in the ED and relate to airway management, fluid
resuscitation, and chest injury management. The only audit filter
directly associated with airway management was "GCS under 9 and not
intubated" which performed poorly in our case. This could suggest that
an alternative audit filter related to airway management could
potentially perform differently.

The sole purpose of audit filters is to elevate the quality of care.
Audit filters should therefore reflect the current quality gap. However,
once quality is improved, these audit filters loose their purpose which
can result in their performance dropping. This could suggest that rather
than the audit filters performing poorly, they have actually fulfilled
their purpose. It is therefore crucial to continuously update the audit
filters. \## Generalizability

Given the variability in screening for OFI, what an OFI consists of and
what specific audit filters are used at a given trauma centre the
results will vary too. Likewise, audit filters are used to fill quality
gaps and due to individual trauma centre suffering from unique quality
gaps, there will not be a golden standard set of audit filters that work
everywhere.

Audit filters and quality improvement programs are widely used and the
findings in this study suggest that some audit filters perform better
than others. The audit filters with performance exceeding coincidence
(AUC = 0.5) might be a beneficial add-on to an already existing quality
improvement programme or work as a guidance in fine-tuning audit filters
already in use to suit the hospitals' quality gaps. Due to the study
being conducted on a single, trauma level I centre in Stockholm it is
unjust to assume applicability of these findings at other hospitals in
other countries facing different challenges in trauma care. It is
already suggested that the efficacy of audit filters vary depending on
country. Low and middle income countries seem to benefit from certain
audit filters that high income countries do not and vice versa
[@berg2022perceived].

## Strengths

As stated in the method section, some audit filters were manually
created from registry data instead of relying on the selection process
by the specialised nurse. This, in order to minimise misclassification
bias.

Furthermore, there has been very limited high quality evidence
supporting the efficacy of audit filters [@evans2009audit]. This study
adds to the evidence suggesting that some audit filters perform better
than others.

Previous studies focus primarily on audit filters' ability to reduce
mortality. In this study the morbidity of patient cases is also taken
into account using OFIs as an umbrella term consisting of everything
from bad documentation to preventable death.

From 2017 all trauma patients admitted to Karolinska University Hospital
were included in the registry reducing the risk of bias in inclusion of
patient cases.

## Limitations

Opportunity for improvement, while dichotomously defined, includes a
diverse set of outcomes. This makes it difficult to create audit filters
that do well in predicting OFIs.

The current screening system for OFIs might also introduce bias since
the filters would favour the identification of some, but not all,
errors.

The way the performance of audit filters were individually calculated,
as stated in the method section, was to look at every patient case with
or without an OFI and simultaneously flagged or not flagged with the
audit filter. Consequently, even if an audit filter demonstrated a true
positive correlation with an OFI, instances where the same patient case
had multiple audit filters introduce a degree of uncertainty regarding
which specific audit filter accurately predicted the outcome.

The number of missing values varied greatly among the audit filters. As
stated above, missing values from each audit filter was coded as "not
flagged" by an audit filter. This means that in some instances where
patient cases not flagged, might actually meet the audit filter
criteria, but for some reason not documented. This reduces the validity
of some audit filters who displayed high amounts of missing data.

## Future Studies

Given the expanding field of artificial intelligence and machine
learning, newer studies should focus on using these tools to screen for
patient cases at risk of an OFI. This will reduce misclassification and
ensure constant and continued good performance due to its ability to
learn from the current context of trauma care. This approach is new but
not unexplored. A preprint is currently available by Attergrim et. al
[@attergrim2023predicting] exploring the idea of machine learning models
as screening tool for potential OFIs. They found that machine learning
models outperformed the currently used audit filters in predicting OFIs.

If the use of audit filter is still considered, it is important to
ensure its performance. Future studies could look at subgroups of OFIs,
and which audit filter responds to that subgroup. Some audit filters
might perform well in predicting a subgroup of OFIs but fail to predict
others.

## Health Equity

It could be of interest to study the occurrence of OFIs among
marginalised groups of people and women in order to assess whether some
groups tend to fall out from the screening method or are at a higher
rate falsely labelled with or without OFIs.

Since there is a gender disparity between the number of trauma patients,
looking further into whether gender disparities in OFIs could reveal
gender discrimination in the screening and labelling of OFIs.

# Conclusion

Audit filters as a tool in quality improvement programmes are widely
used. Strong evidence supporting their efficacy in quality improvement
is scarse, but seem to depend on where and when they are used, in what
context and for what purpose. In this study, no audit filter seemed to
show good performance in predicting OFIs with AUC ranging from 0.5 to
0.63. These results highlight the importance of either continually
evaluating the relevancy of certain audit filters. Audit filters should
reflect the current gap in trauma care quality. This process takes time
and resources and using an alternative to audit filters is recommended.

# Acknowledgements

I want to thank my supervisor Martin Gerdin Wärnberg and my
co-supervisors Jonatan Attergrim and Kelvin Szolnoky.

# Tables

# References

::: {#refs}
:::

#Appendices
